---
title: "NCSS_predict_02_23_26"
format: html
---
###  Packages

```{r}
library(tidyverse)
library(tidymodels)
library(recipes)
library(uwot)
library(forcats)
library(beepr)
library(yardstick)
library(dbscan)
library(rlang)

```
### Data prep and split
```{r}
#load dataset:
#150cm:
pedon_pack_150_p <- read.csv("../outputs/pedon_pack_150_03_05_2026.csv")

set.seed(1)

df <- pedon_pack_150_p %>%
  mutate(
    peiid = as.character(peiid),
    taxorder = factor(taxorder)
  )

# lump rare orders to avoid dropped-level issues in resampling
min_n <- 15
df <- df %>%
  mutate(taxorder = fct_lump_min(taxorder, min = min_n, other_level = "Other"))
#there is only 4 lol...
count(df, taxorder, sort = TRUE)

df <- df %>% filter(taxorder != "Other")
df <- df %>% droplevels()
count(df, taxorder, sort = TRUE)

split <- initial_split(df, prop = 0.8, strata = taxorder)
train <- training(split)
test  <- testing(split)
```

#Taxorder:

### Recipe + TAXORDER UMAP: 

```{r}
rec <- recipe(taxorder ~ ., data = train) %>%
  update_role(peiid, new_role = "id") %>%
  step_rm(any_of(c(
    "taxonname_clean","taxonname_key","taxclname",
    "taxsuborder","taxgrtgroup","taxpartsize",
    "taxsubgrp","taxsubgrp_clean",
    "taxsubgrp_mod1","taxsubgrp_mod2","taxsubgrp_mod3",
    "taxsubgrp_mod_group","taxgrtgroup_mod_group","taxsuborder_mod_group"
  ))) %>%
  step_rm(starts_with("mod_"), starts_with("is_")) %>%

  step_string2factor(all_nominal_predictors()) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%

  # NA indicators ONLY for CEC-related columns
  step_indicate_na(all_of(c(
    "cec7_w",
    "pct_hz_with_cec7",
    "min_cec_per_kg_clay_0_150",
    "median_cec_per_kg_clay_bt_zone"
  )), prefix = "na2_") %>%

  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%

  step_impute_median(all_numeric_predictors())  %>%
  step_mutate_at(
    all_of(c(
      "sandy_mantle_cm",
      "plinthite_thickness_cm","plinthite_overlap_0_150_cm",
      "bt_thickness_cm","argillic_thickness_cm","argillic_contiguous_max_cm"
    )),
    fn = ~ pmax(.x, 0)
  ) %>%
  step_log(all_of(c(
    "sandy_mantle_cm",
    "plinthite_thickness_cm","plinthite_overlap_0_150_cm",
    "bt_thickness_cm","argillic_thickness_cm","argillic_contiguous_max_cm"
  )), offset = 1) %>%
  # log BEFORE normalize
  step_log(all_of(c(
    "sandy_mantle_cm",
    "plinthite_thickness_cm","plinthite_overlap_0_150_cm",
    "bt_thickness_cm","argillic_thickness_cm","argillic_contiguous_max_cm"
  )), offset = 1) %>%

  step_normalize(all_numeric_predictors())

prep_rec <- prep(rec)

x_train <- bake(prep_rec, new_data = train) %>% select(-taxorder, -peiid)
x_test  <- bake(prep_rec, new_data = test)  %>% select(-taxorder, -peiid)

# convert logical predictors to 0/1
x_train <- x_train %>% mutate(across(where(is.logical), as.integer))
x_test  <- x_test  %>% mutate(across(where(is.logical), as.integer))

# safety: ensure all numeric
stopifnot(all(vapply(x_train, is.numeric, logical(1))))


#HDBSCAN PREP:
library(dbscan)
library(rlang)

# ------------------------------------------------------------
# Add HDBSCAN-derived features to a train/test UMAP pair
# ------------------------------------------------------------
add_hdbscan_features <- function(train_umap, test_umap, outcome, minPts = 10) {
  
  x_train <- train_umap %>% select(-all_of(outcome))
  x_test  <- test_umap  %>% select(-all_of(outcome))
  
  hdb_fit <- dbscan::hdbscan(
    as.matrix(x_train),
    minPts = minPts
  )
  
  test_cluster <- predict(
    hdb_fit,
    newdata = as.matrix(x_test),
    data    = as.matrix(x_train)
  )
  
  levs <- sort(unique(c(hdb_fit$cluster, test_cluster)))
  
  train_hdb <- train_umap %>%
    mutate(
      hdb_cluster = factor(hdb_fit$cluster, levels = levs),
      hdb_noise   = factor(if_else(hdb_fit$cluster == 0, "noise", "cluster"),
                           levels = c("cluster", "noise"))
    )
  
  test_hdb <- test_umap %>%
    mutate(
      hdb_cluster = factor(test_cluster, levels = levs),
      hdb_noise   = factor(if_else(test_cluster == 0, "noise", "cluster"),
                           levels = c("cluster", "noise"))
    )
  
  list(
    train = train_hdb,
    test  = test_hdb,
    fit   = hdb_fit
  )
}

# ------------------------------------------------------------
# Fit tuned RF on any train/test set with a single outcome
# ------------------------------------------------------------
fit_tuned_rf_classifier <- function(train_dat,
                                    test_dat,
                                    outcome,
                                    trees = 1500,
                                    metric_select = "bal_accuracy",
                                    grid_levels = 4,
                                    mtry_max = NULL,
                                    min_n_max = 20,
                                    seed_cv = 1) {
  
  outcome_sym <- rlang::sym(outcome)
  
  n_pred <- ncol(train_dat) - 1
  if (is.null(mtry_max)) {
    mtry_max <- min(10, n_pred)
  } else {
    mtry_max <- min(mtry_max, n_pred)
  }
  
  if (mtry_max < 2) stop("Not enough predictors for tuning mtry.")
  
  min_class_train <- train_dat %>%
    count(!!outcome_sym) %>%
    summarise(min_n = min(n)) %>%
    pull(min_n)
  
  v <- min(5, min_class_train)
  if (v < 2) stop("At least one class has <2 training samples.")
  
  set.seed(seed_cv)
  folds <- vfold_cv(train_dat, v = v, strata = !!outcome_sym)
  
  rf_spec <- rand_forest(
    trees = trees,
    mtry  = tune(),
    min_n = tune()
  ) %>%
    set_engine("ranger", probability = TRUE, importance = "permutation") %>%
    set_mode("classification")
  
  wf <- workflow() %>%
    add_model(rf_spec) %>%
    add_formula(as.formula(paste(outcome, "~ .")))
  
  grid <- grid_regular(
    mtry(range = c(2, mtry_max)),
    min_n(range = c(2, min_n_max)),
    levels = grid_levels
  )
  
  tuned <- tune_grid(
    wf,
    resamples = folds,
    grid = grid,
    metrics = metric_set(accuracy, kap, bal_accuracy)
  )
  
  best <- select_best(tuned, metric = metric_select)
  
  final_fit <- finalize_workflow(wf, best) %>%
    fit(train_dat)
  
  pred <- predict(final_fit, test_dat, type = "prob") %>%
    bind_cols(predict(final_fit, test_dat, type = "class")) %>%
    bind_cols(test_dat %>% select(all_of(outcome)))
  
  list(
    tuned = tuned,
    best  = best,
    fit   = final_fit,
    pred  = pred
  )
}

# ------------------------------------------------------------
# Compare baseline vs HDBSCAN-augmented RF
# ------------------------------------------------------------
metric_compare_tbl <- function(pred_base, pred_hdb, outcome) {
  outcome_sym <- rlang::sym(outcome)
  
  base_tbl <- bind_rows(
    metric_set(accuracy, kap, bal_accuracy)(
      pred_base, truth = !!outcome_sym, estimate = .pred_class
    ),
    precision(
      pred_base, truth = !!outcome_sym, estimate = .pred_class, estimator = "macro"
    ),
    recall(
      pred_base, truth = !!outcome_sym, estimate = .pred_class, estimator = "macro"
    ),
    f_meas(
      pred_base, truth = !!outcome_sym, estimate = .pred_class, estimator = "macro"
    )
  ) %>%
    mutate(model = "baseline", .before = 1)
  
  hdb_tbl <- bind_rows(
    metric_set(accuracy, kap, bal_accuracy)(
      pred_hdb, truth = !!outcome_sym, estimate = .pred_class
    ),
    precision(
      pred_hdb, truth = !!outcome_sym, estimate = .pred_class, estimator = "macro"
    ),
    recall(
      pred_hdb, truth = !!outcome_sym, estimate = .pred_class, estimator = "macro"
    ),
    f_meas(
      pred_hdb, truth = !!outcome_sym, estimate = .pred_class, estimator = "macro"
    )
  ) %>%
    mutate(model = "hdbscan", .before = 1)
  
  bind_rows(base_tbl, hdb_tbl)
}

# ------------------------------------------------------------
# Confidence summary helper
# ------------------------------------------------------------
confidence_summary_tbl <- function(pred_tbl, model_name = "model") {
  pred_tbl %>%
    rowwise() %>%
    mutate(
      probs  = list(c_across(starts_with(".pred_") & !matches("^\\.pred_class$"))),
      top1   = max(probs),
      top2   = ifelse(length(probs) > 1, sort(probs, decreasing = TRUE)[2], NA_real_),
      margin = ifelse(length(probs) > 1, top1 - top2, NA_real_)
    ) %>%
    ungroup() %>%
    summarise(
      model          = model_name,
      mean_top1      = mean(top1, na.rm = TRUE),
      pct_low_top1   = mean(top1 < 0.5, na.rm = TRUE),
      mean_margin    = mean(margin, na.rm = TRUE),
      pct_low_margin = mean(margin < 0.15, na.rm = TRUE)
    )
}
```

### TAXORDER UMAP and DBSCAN

```{r}
# UMAP can't take NA/Inf; enforce finite numeric matrix
x_train_umap <- x_train %>%
  mutate(across(everything(), ~ ifelse(is.finite(.x), .x, NA_real_)))
x_test_umap <- x_test %>%
  mutate(across(everything(), ~ ifelse(is.finite(.x), .x, NA_real_)))

# replace any remaining NA with 0 (UMAP-safe)
x_train_umap[is.na(x_train_umap)] <- 0
x_test_umap[is.na(x_test_umap)] <- 0

set.seed(2)
umap_fit <- uwot::umap(
  as.matrix(x_train_umap),
  n_components = 10,
  n_neighbors = 15,
  min_dist = 0.1,
  metric = "euclidean",
  ret_model = TRUE
)

z_test <- as.data.frame(uwot::umap_transform(as.matrix(x_test_umap), umap_fit))
z_train <- as.data.frame(umap_fit$embedding)
names(z_train) <- paste0("U", seq_len(ncol(z_train)))
names(z_test) <- names(z_train)

train_umap <- bind_cols(z_train, taxorder = train$taxorder)
test_umap  <- bind_cols(z_test,  taxorder = test$taxorder)



#UMap is a non-linear dimensionality reduction technique that can help capture complex relationships in the data. By reducing the dimensionality of the predictor space, it can make it easier for models like random forests to learn patterns, especially when dealing with many predictors or complex interactions. However, it’s important to note that UMAP is unsupervised and may not always improve performance, so it’s worth comparing results with and without UMAP as part of your modeling workflow.


```

### TAXORDER HDBSCAN

```{r}
#HDBSCAN:


set.seed(4)

# fit HDBSCAN on training UMAP coordinates only
hdb_fit <- dbscan::hdbscan(
  as.matrix(z_train),
  minPts = 10
)

# predict cluster IDs for test UMAP coordinates
# 0 = noise
test_hdb_cluster <- predict(
  hdb_fit,
  newdata = as.matrix(z_test),
  data    = as.matrix(z_train)
)

# make sure train/test use the same factor levels
hdb_levels <- sort(unique(c(hdb_fit$cluster, test_hdb_cluster)))

train_umap_hdb <- train_umap %>%
  mutate(
    hdb_cluster = factor(hdb_fit$cluster, levels = hdb_levels),
    hdb_noise   = factor(if_else(hdb_fit$cluster == 0, "noise", "cluster"),
                         levels = c("cluster", "noise"))
  )

test_umap_hdb <- test_umap %>%
  mutate(
    hdb_cluster = factor(test_hdb_cluster, levels = hdb_levels),
    hdb_noise   = factor(if_else(test_hdb_cluster == 0, "noise", "cluster"),
                         levels = c("cluster", "noise"))
  )

count(train_umap_hdb, hdb_cluster, taxorder, sort = TRUE)
count(test_umap_hdb,  hdb_cluster, taxorder, sort = TRUE)

rf_spec <- rand_forest(trees = 1000) %>%
  set_engine("ranger", probability = TRUE, importance = "permutation") %>%
  set_mode("classification")

rf_fit_hdb <- workflow() %>%
  add_model(rf_spec) %>%
  add_formula(taxorder ~ .) %>%
  fit(train_umap_hdb)

pred_class_hdb <- predict(rf_fit_hdb, test_umap_hdb, type = "class") %>%
  bind_cols(test_umap_hdb %>% select(taxorder))

conf_mat(pred_class_hdb, truth = taxorder, estimate = .pred_class)
accuracy(pred_class_hdb, truth = taxorder, estimate = .pred_class)

rf_spec_tune <- rand_forest(
  trees = 1000,
  mtry  = tune(),
  min_n = tune()
) %>%
  set_engine("ranger", probability = TRUE, importance = "permutation") %>%
  set_mode("classification")

wf_tune_hdb <- workflow() %>%
  add_model(rf_spec_tune) %>%
  add_formula(taxorder ~ .)

set.seed(3)
folds_hdb <- vfold_cv(train_umap_hdb, v = 5, strata = taxorder)

p <- ncol(train_umap_hdb) - 1   # minus outcome column

grid_hdb <- grid_regular(
  mtry(range = c(2, min(12, p))),
  min_n(range = c(2, 25)),
  levels = 5
)

tuned_hdb <- tune_grid(
  wf_tune_hdb,
  resamples = folds_hdb,
  grid = grid_hdb,
  metrics = metric_set(accuracy, kap)
)

best_hdb <- select_best(tuned_hdb, metric = "kap")

rf_final_hdb <- finalize_workflow(wf_tune_hdb, best_hdb) %>%
  fit(train_umap_hdb)

pred_hdb <- predict(rf_final_hdb, test_umap_hdb, type = "prob") %>%
  bind_cols(predict(rf_final_hdb, test_umap_hdb, type = "class")) %>%
  bind_cols(test_umap_hdb %>% select(taxorder))

conf_mat(pred_hdb, truth = taxorder, estimate = .pred_class)
accuracy(pred_hdb, truth = taxorder, estimate = .pred_class)
kap(pred_hdb, truth = taxorder, estimate = .pred_class)
bal_accuracy(pred_hdb, truth = taxorder, estimate = .pred_class)

pred_hdb %>%
  precision(truth = taxorder, estimate = .pred_class, estimator = "macro")

pred_hdb %>%
  recall(truth = taxorder, estimate = .pred_class, estimator = "macro")

pred_hdb %>%
  f_meas(truth = taxorder, estimate = .pred_class, estimator = "macro")

pred_conf_hdb <- pred_hdb %>%
  rowwise() %>%
  mutate(
    max_prob = max(c_across(starts_with(".pred_") & !matches("^\\.pred_class$")))
  ) %>%
  ungroup() %>%
  summarise(
    mean_max_prob = mean(max_prob),
    pct_low_conf  = mean(max_prob < 0.5)
  )

pred_conf_hdb


```

### TAXORDER RF on UMAP only (no HDBSCAN features)

```{r}
rf_spec <- rand_forest(trees = 1000) %>%
  set_engine("ranger", probability = TRUE, importance = "permutation") %>%
  set_mode("classification")

rf_fit <- workflow() %>%
  add_model(rf_spec) %>%
  add_formula(taxorder ~ .) %>%
  fit(train_umap)

pred_class <- predict(rf_fit, test_umap, type = "class") %>%
  bind_cols(test_umap %>% select(taxorder))

conf_mat(pred_class, truth = taxorder, estimate = .pred_class)
accuracy(pred_class, truth = taxorder, estimate = .pred_class)
```

### TAXORDER RF with UMAP + HDBSCAN features (tuned)

```{r}
rf_spec_tune <- rand_forest(trees = 1000, mtry = tune(), min_n = tune()) %>%
  set_engine("ranger", probability = TRUE, importance = "permutation") %>%
  set_mode("classification")

wf_tune <- workflow() %>%
  add_model(rf_spec_tune) %>%
  add_formula(taxorder ~ .)

set.seed(4)
folds <- vfold_cv(train_umap, v = 5, strata = taxorder)

grid <- grid_regular(
  mtry(range = c(2, 10)),
  min_n(range = c(2, 25)),
  levels = 5
)

tuned <- tune_grid(
  wf_tune, resamples = folds, grid = grid,
  metrics = metric_set(accuracy, kap)
)
beepr::beep("wilhelm")
best <- select_best(tuned, metric = "kap")
rf_final <- finalize_workflow(wf_tune, best) %>%
  fit(train_umap)
pred <- predict(rf_final, test_umap, type = "prob") %>%
  bind_cols(predict(rf_final, test_umap, type = "class")) %>%
  bind_cols(test_umap %>% select(taxorder))

conf_mat(pred, truth = taxorder, estimate = .pred_class)
kap(pred, truth = taxorder, estimate = .pred_class)

pred_conf <- pred %>%
  rowwise() %>%
  mutate(
    max_prob = max(c_across(starts_with(".pred_") & !matches("^\\.pred_class$")))
  ) %>%
  ungroup() %>%
  summarise(
    mean_max_prob = mean(max_prob),
    pct_low_conf  = mean(max_prob < 0.5)
  )
pred_conf

```

### TAXORDER Model eval

```{r}
conf_mat(pred, truth = taxorder, estimate = .pred_class)

# per-class stats (super important if classes are imbal)
pred %>% yardstick::precision(truth = taxorder, estimate = .pred_class, estimator = "macro")
pred %>% yardstick::recall(truth = taxorder, estimate = .pred_class, estimator = "macro")
pred %>% yardstick::f_meas(truth = taxorder, estimate = .pred_class, estimator = "macro")
```
```{r}
pred <- pred %>%
  rowwise() %>%
  mutate(max_prob = max(c_across(where(is.numeric) & starts_with(".pred_")))) %>%
  ungroup()

pred %>%
  mutate(low_conf = max_prob < 0.5) %>%
  count(taxorder, low_conf, sort = TRUE)

pred_class <- pred %>%
  select(taxorder, .pred_class)

yardstick::bal_accuracy(pred_class, truth = taxorder, estimate = .pred_class)

f_meas(pred_class, truth = taxorder, estimate = .pred_class, estimator = "macro")
recall(pred_class, truth = taxorder, estimate = .pred_class, estimator = "macro")
precision(pred_class, truth = taxorder, estimate = .pred_class, estimator = "macro")
```

### TAXORDER performance

```{r}
# 1) Class counts (train + test)
cat("\n=== FULL DATA COUNTS ===\n")
print(count(df, taxorder, sort = TRUE))

cat("\n=== TRAIN COUNTS ===\n")
print(count(train, taxorder, sort = TRUE))

cat("\n=== TEST COUNTS ===\n")
print(count(test, taxorder, sort = TRUE))

# 2) Confusion matrix (full)
cat("\n=== CONFUSION MATRIX ===\n")
print(conf_mat(pred, truth = taxorder, estimate = .pred_class))

# 3) Key metrics (overall + macro)
cat("\n=== METRICS (overall) ===\n")
print(metric_set(accuracy, kap, bal_accuracy)(pred, truth = taxorder, estimate = .pred_class))

cat("\n=== MACRO METRICS ===\n")
print(precision(pred, truth = taxorder, estimate = .pred_class, estimator = "macro"))
print(recall(pred, truth = taxorder, estimate = .pred_class, estimator = "macro"))
print(f_meas(pred, truth = taxorder, estimate = .pred_class, estimator = "macro"))

# 4) Per-class performance (this is the most diagnostic)
cat("\n=== PER-CLASS METRICS ===\n")
print(precision(pred, truth = taxorder, estimate = .pred_class, estimator = "macro_weighted")) # optional
print(recall(pred, truth = taxorder, estimate = .pred_class, estimator = "macro_weighted"))   # optional


cat("\n=== BY-CLASS SUMMARY ===\n")
print(summary(conf_mat(pred, truth = taxorder, estimate = .pred_class)))

#So: the model works, but it’s blind to Alfisols.
```

```{r}
#hdbscan vs raw umap:
cat("\n=== BASELINE RF (UMAP only) ===\n")
print(metric_set(accuracy, kap, bal_accuracy)(pred, truth = taxorder, estimate = .pred_class))
print(precision(pred, truth = taxorder, estimate = .pred_class, estimator = "macro"))
print(recall(pred, truth = taxorder, estimate = .pred_class, estimator = "macro"))
print(f_meas(pred, truth = taxorder, estimate = .pred_class, estimator = "macro"))

cat("\n=== RF + HDBSCAN FEATURES ===\n")
print(metric_set(accuracy, kap, bal_accuracy)(pred_hdb, truth = taxorder, estimate = .pred_class))
print(precision(pred_hdb, truth = taxorder, estimate = .pred_class, estimator = "macro"))
print(recall(pred_hdb, truth = taxorder, estimate = .pred_class, estimator = "macro"))
print(f_meas(pred_hdb, truth = taxorder, estimate = .pred_class, estimator = "macro"))
```

### Taxsuborder:

```{r}
# ============================================================
# TAXSUBORDER MODEL (taxsuborder)
# ============================================================

set.seed(51)

min_n_so <- 15   # bump to 20 if CV gets unstable

df_so <- pedon_pack_150_p %>%
  mutate(
    peiid = as.character(peiid),
    taxsuborder = str_squish(as.character(taxsuborder)),
    taxsuborder = na_if(taxsuborder, "")
  ) %>%
  filter(!is.na(taxsuborder)) %>%
  mutate(taxsuborder = factor(taxsuborder))

# lump rare classes
df_so <- df_so %>%
  mutate(
    taxsuborder = fct_lump_min(
      taxsuborder,
      min = min_n_so,
      other_level = "Other"
    )
  ) %>%
  filter(taxsuborder != "Other") %>%
  droplevels()

cat("\n=== TAXSUBORDER COUNTS ===\n")
print(count(df_so, taxsuborder, sort = TRUE))

split_so <- initial_split(df_so, prop = 0.8, strata = taxsuborder)
train_so <- training(split_so)
test_so  <- testing(split_so)
rec_so <- recipe(taxsuborder ~ ., data = train_so) %>%
  update_role(peiid, new_role = "id") %>%
  step_rm(any_of(c(
    "taxonname_clean", "taxonname_key", "taxclname",
    "taxorder",       # important: remove parent class
    "taxgrtgroup",
    "taxsubgrp_mod1", "taxsubgrp_mod2", "taxsubgrp_mod3",
    "taxpartsize"
  ))) %>%
  step_string2factor(all_nominal_predictors()) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_zv(all_predictors()) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_normalize(all_numeric_predictors())

prep_rec_so <- prep(rec_so)

x_train_so <- bake(prep_rec_so, new_data = train_so) %>%
  select(-taxsuborder, -peiid)

x_test_so <- bake(prep_rec_so, new_data = test_so) %>%
  select(-taxsuborder, -peiid)

x_train_so <- x_train_so %>% mutate(across(where(is.logical), as.integer))
x_test_so  <- x_test_so  %>% mutate(across(where(is.logical), as.integer))

stopifnot(all(vapply(x_train_so, is.numeric, logical(1))))

cat("\nTrain rows:", nrow(x_train_so), "\n")
cat("Train cols:", ncol(x_train_so), "\n")
set.seed(52)

n_neighbors_so  <- min(15, nrow(x_train_so) - 1)
n_components_so <- min(12, ncol(x_train_so))

umap_fit_so <- uwot::umap(
  as.matrix(x_train_so),
  n_components = n_components_so,
  n_neighbors = n_neighbors_so,
  min_dist = 0.1,
  metric = "euclidean",
  ret_model = TRUE
)

z_train_so <- as.data.frame(umap_fit_so$embedding)
names(z_train_so) <- paste0("U", seq_len(ncol(z_train_so)))

z_test_so <- as.data.frame(uwot::umap_transform(as.matrix(x_test_so), umap_fit_so))
names(z_test_so) <- names(z_train_so)

train_umap_so <- bind_cols(z_train_so, taxsuborder = train_so$taxsuborder)
test_umap_so  <- bind_cols(z_test_so,  taxsuborder = test_so$taxsuborder)
# ------------------------------------------------------------
# TAXSUBORDER BASELINE RF
# ------------------------------------------------------------
rf_so_base <- fit_tuned_rf_classifier(
  train_dat      = train_umap_so,
  test_dat       = test_umap_so,
  outcome        = "taxsuborder",
  trees          = 1500,
  metric_select  = "bal_accuracy",
  grid_levels    = 5,
  mtry_max       = min(12, ncol(train_umap_so) - 1),
  min_n_max      = 25,
  seed_cv        = 53
)

pred_so <- rf_so_base$pred

cat("\n=== TAXSUBORDER BASELINE METRICS ===\n")
print(metric_set(accuracy, kap, bal_accuracy)(
  pred_so,
  truth = taxsuborder,
  estimate = .pred_class
))

print(precision(
  pred_so,
  truth = taxsuborder,
  estimate = .pred_class,
  estimator = "macro"
))

print(recall(
  pred_so,
  truth = taxsuborder,
  estimate = .pred_class,
  estimator = "macro"
))

print(f_meas(
  pred_so,
  truth = taxsuborder,
  estimate = .pred_class,
  estimator = "macro"
))

cat("\n=== TAXSUBORDER BASELINE CONFUSION MATRIX ===\n")
print(conf_mat(pred_so, truth = taxsuborder, estimate = .pred_class))

cat("\n=== TAXSUBORDER BASELINE CONFUSION MATRIX SUMMARY ===\n")
print(summary(conf_mat(pred_so, truth = taxsuborder, estimate = .pred_class)))

cat("\n=== TAXSUBORDER BASELINE CONFIDENCE ===\n")
print(confidence_summary_tbl(pred_so, "baseline"))
# ------------------------------------------------------------
# TAXSUBORDER HDBSCAN + RF
# ------------------------------------------------------------
hdb_so <- add_hdbscan_features(
  train_umap = train_umap_so,
  test_umap  = test_umap_so,
  outcome    = "taxsuborder",
  minPts     = 10
)

cat("\n=== TAXSUBORDER HDBSCAN CLUSTER COUNTS (TRAIN) ===\n")
print(count(hdb_so$train, hdb_cluster, taxsuborder, sort = TRUE))

rf_so_hdb <- fit_tuned_rf_classifier(
  train_dat      = hdb_so$train,
  test_dat       = hdb_so$test,
  outcome        = "taxsuborder",
  trees          = 1500,
  metric_select  = "bal_accuracy",
  grid_levels    = 5,
  mtry_max       = min(12, ncol(hdb_so$train) - 1),
  min_n_max      = 25,
  seed_cv        = 54
)

pred_so_hdb <- rf_so_hdb$pred

cat("\n=== TAXSUBORDER HDBSCAN METRICS ===\n")
print(metric_set(accuracy, kap, bal_accuracy)(
  pred_so_hdb,
  truth = taxsuborder,
  estimate = .pred_class
))

print(precision(
  pred_so_hdb,
  truth = taxsuborder,
  estimate = .pred_class,
  estimator = "macro"
))

print(recall(
  pred_so_hdb,
  truth = taxsuborder,
  estimate = .pred_class,
  estimator = "macro"
))

print(f_meas(
  pred_so_hdb,
  truth = taxsuborder,
  estimate = .pred_class,
  estimator = "macro"
))

cat("\n=== TAXSUBORDER HDBSCAN CONFUSION MATRIX ===\n")
print(conf_mat(pred_so_hdb, truth = taxsuborder, estimate = .pred_class))

cat("\n=== TAXSUBORDER HDBSCAN CONFUSION MATRIX SUMMARY ===\n")
print(summary(conf_mat(pred_so_hdb, truth = taxsuborder, estimate = .pred_class)))

cat("\n=== TAXSUBORDER HDBSCAN CONFIDENCE ===\n")
print(confidence_summary_tbl(pred_so_hdb, "hdbscan"))
cat("\n=== TAXSUBORDER: BASELINE VS HDBSCAN ===\n")
print(metric_compare_tbl(pred_so, pred_so_hdb, outcome = "taxsuborder"))

cat("\n=== TAXSUBORDER CONFIDENCE: BASELINE VS HDBSCAN ===\n")
print(bind_rows(
  confidence_summary_tbl(pred_so, "baseline"),
  confidence_summary_tbl(pred_so_hdb, "hdbscan")
))

cat("\n=== TAXSUBORDER TRUTH VS PREDICTED COUNTS (BASELINE) ===\n")
print(pred_so %>% count(taxsuborder, .pred_class, sort = TRUE))

cat("\n=== TAXSUBORDER TRUTH VS PREDICTED COUNTS (HDBSCAN) ===\n")
print(pred_so_hdb %>% count(taxsuborder, .pred_class, sort = TRUE))
```

#GreatGroup:



```{r}
# ============================================================
# GREAT GROUP MODEL (taxgrtgroup)
# ============================================================

set.seed(11)

df_gg <- pedon_pack_150_p %>%
  mutate(
    peiid = as.character(peiid),
    taxgrtgroup = factor(taxgrtgroup)
  ) %>%
  filter(!is.na(taxgrtgroup), taxgrtgroup != "") %>%
  droplevels()

# Lump rare great groups (you almost always need this)
# Start with 10–20; increase if you still see dropped-level warnings in CV
min_n_gg <- 20
df_gg <- df_gg %>%
  mutate(taxgrtgroup = fct_lump_min(taxgrtgroup, min = min_n_gg, other_level = "Other")) %>%
  filter(taxgrtgroup != "Other") %>%   # you can keep Other if you want, but it’s often messy
  droplevels()

cat("\n=== GREAT GROUP COUNTS ===\n")
print(count(df_gg, taxgrtgroup, sort = TRUE))

split_gg <- initial_split(df_gg, prop = 0.8, strata = taxgrtgroup)
train_gg <- training(split_gg)
test_gg  <- testing(split_gg)

# --- recipe: drop taxonomy leaks (everything except taxgrtgroup)
rec_gg <- recipe(taxgrtgroup ~ ., data = train_gg) %>%
  update_role(peiid, new_role = "id") %>%
  step_rm(any_of(c(
    "taxonname_clean",
    "taxorder", "taxsuborder",
    "taxsubgrp_mod1", "taxsubgrp_mod2", "taxsubgrp_mod3",
    "taxpartsize"
  ))) %>%
  step_zv(all_predictors()) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_normalize(all_numeric_predictors())

prep_rec_gg <- prep(rec_gg)

x_train_gg <- bake(prep_rec_gg, new_data = train_gg) %>% select(-taxgrtgroup, -peiid)
x_test_gg  <- bake(prep_rec_gg, new_data = test_gg)  %>% select(-taxgrtgroup, -peiid)

x_train_gg <- x_train_gg %>% mutate(across(where(is.logical), as.integer))
x_test_gg  <- x_test_gg  %>% mutate(across(where(is.logical), as.integer))

stopifnot(all(vapply(x_train_gg, is.numeric, logical(1))))

# --- UMAP
set.seed(12)
umap_fit_gg <- uwot::umap(
  as.matrix(x_train_gg),
  n_components = 15,     # often helps for many-class problems
  n_neighbors = 15,
  min_dist = 0.1,
  metric = "euclidean",
  ret_model = TRUE
)

z_train_gg <- as.data.frame(umap_fit_gg$embedding)
names(z_train_gg) <- paste0("U", seq_len(ncol(z_train_gg)))

z_test_gg <- as.data.frame(uwot::umap_transform(as.matrix(x_test_gg), umap_fit_gg))
names(z_test_gg) <- names(z_train_gg)

train_umap_gg <- bind_cols(z_train_gg, taxgrtgroup = train_gg$taxgrtgroup)
test_umap_gg  <- bind_cols(z_test_gg,  taxgrtgroup = test_gg$taxgrtgroup)

# --- RF tune (recommended for GG)
rf_spec_tune_gg <- rand_forest(trees = 1500, mtry = tune(), min_n = tune()) %>%
  set_engine("ranger", probability = TRUE, importance = "permutation") %>%
  set_mode("classification")

wf_tune_gg <- workflow() %>%
  add_model(rf_spec_tune_gg) %>%
  add_formula(taxgrtgroup ~ .)

set.seed(13)
folds_gg <- vfold_cv(train_umap_gg, v = 5, strata = taxgrtgroup)

# mtry upper bound should not exceed number of UMAP components (here 15)
grid_gg <- grid_regular(
  mtry(range = c(2, 15)),
  min_n(range = c(2, 30)),
  levels = 5
)

tuned_gg <- tune_grid(
  wf_tune_gg, resamples = folds_gg, grid = grid_gg,
  metrics = metric_set(bal_accuracy, kap)
)
beepr::beep("wilhelm")

best_gg <- select_best(tuned_gg, metric = "bal_accuracy")

rf_final_gg <- finalize_workflow(wf_tune_gg, best_gg) %>%
  fit(train_umap_gg)

pred_gg <- predict(rf_final_gg, test_umap_gg, type = "prob") %>%
  bind_cols(predict(rf_final_gg, test_umap_gg, type = "class")) %>%
  bind_cols(test_umap_gg %>% select(taxgrtgroup))
pred_gg <- pred_gg %>%
  rowwise() %>%
  mutate(
    probs = list(c_across(where(is.numeric) & starts_with(".pred_"))),
    top1   = max(probs),
    top2   = sort(probs, decreasing = TRUE)[2],
    margin = top1 - top2
  ) %>%
  ungroup()

pred_gg %>%
  summarise(
    mean_top1 = mean(top1),
    pct_low_top1 = mean(top1 < 0.5),
    mean_margin = mean(margin),
    pct_low_margin = mean(margin < 0.15)
  )
cat("\n=== GREAT GROUP CONFUSION MATRIX ===\n")
print(conf_mat(pred_gg, truth = taxgrtgroup, estimate = .pred_class))

cat("\n=== GREAT GROUP METRICS ===\n")
print(metric_set(accuracy, kap, bal_accuracy)(pred_gg, truth = taxgrtgroup, estimate = .pred_class))
print(f_meas(pred_gg, truth = taxgrtgroup, estimate = .pred_class, estimator = "macro"))

# Confidence summary
pred_conf_gg <- pred_gg %>%
  rowwise() %>%
  mutate(max_prob = max(c_across(where(is.numeric) & starts_with(".pred_")))) %>%
  ungroup() %>%
  summarise(
    mean_max_prob = mean(max_prob),
    pct_low_conf  = mean(max_prob < 0.5)
  )

# cat("\n=== GREAT GROUP CONFIDENCE ===\n")
# print(pred_conf_gg)

# ------------------------------------------------------------
# GREAT GROUP: HDBSCAN + RF
# ------------------------------------------------------------
hdb_gg <- add_hdbscan_features(
  train_umap = train_umap_gg,
  test_umap  = test_umap_gg,
  outcome    = "taxgrtgroup",
  minPts     = 10
)

cat("\n=== GREAT GROUP HDBSCAN CLUSTER COUNTS (TRAIN) ===\n")
print(count(hdb_gg$train, hdb_cluster, taxgrtgroup, sort = TRUE))

rf_gg_hdb <- fit_tuned_rf_classifier(
  train_dat      = hdb_gg$train,
  test_dat       = hdb_gg$test,
  outcome        = "taxgrtgroup",
  trees          = 1500,
  metric_select  = "bal_accuracy",
  grid_levels    = 5,
  mtry_max       = min(15, ncol(hdb_gg$train) - 1),
  min_n_max      = 30,
  seed_cv        = 113
)

pred_gg_hdb <- rf_gg_hdb$pred

cat("\n=== GREAT GROUP: BASELINE VS HDBSCAN ===\n")
print(metric_compare_tbl(pred_gg, pred_gg_hdb, outcome = "taxgrtgroup"))

cat("\n=== GREAT GROUP CONFIDENCE: BASELINE VS HDBSCAN ===\n")
print(bind_rows(
  confidence_summary_tbl(pred_gg,     "baseline"),
  confidence_summary_tbl(pred_gg_hdb, "hdbscan")
))

cat("\n=== GREAT GROUP HDBSCAN CONFUSION MATRIX ===\n")
print(conf_mat(pred_gg_hdb, truth = taxgrtgroup, estimate = .pred_class))

```
### Hyp. Tuning Test for GG:

```{r}
#hyperparameter tuning for GG
# ------------------------------------------------------------
# Extract one-row metric summary for easy comparison
# ------------------------------------------------------------
metric_extract_tbl <- function(pred_tbl, outcome, model_name) {
  outcome_sym <- rlang::sym(outcome)
  
  met <- metric_set(accuracy, kap, bal_accuracy)(
    pred_tbl,
    truth = !!outcome_sym,
    estimate = .pred_class
  )
  
  f1 <- f_meas(
    pred_tbl,
    truth = !!outcome_sym,
    estimate = .pred_class,
    estimator = "macro"
  )
  
  conf <- confidence_summary_tbl(pred_tbl, model_name)
  
  tibble(
    model = model_name,
    accuracy = met$.estimate[met$.metric == "accuracy"],
    kap = met$.estimate[met$.metric == "kap"],
    bal_accuracy = met$.estimate[met$.metric == "bal_accuracy"],
    f_meas_macro = f1$.estimate,
    mean_top1 = conf$mean_top1,
    pct_low_top1 = conf$pct_low_top1,
    mean_margin = conf$mean_margin,
    pct_low_margin = conf$pct_low_margin
  )
}

# ============================================================
# GREAT GROUP: SMALL TUNING SCREEN FOR UMAP + HDBSCAN + RF
# ============================================================

set.seed(120)

# -----------------------------
# 1) Inner train/validation split from the existing training set
# -----------------------------
split_gg_inner <- initial_split(train_gg, prop = 0.8, strata = taxgrtgroup)
train_gg_inner <- training(split_gg_inner)
val_gg_inner   <- testing(split_gg_inner)

# same preprocessing logic as your main GG workflow
rec_gg_inner <- recipe(taxgrtgroup ~ ., data = train_gg_inner) %>%
  update_role(peiid, new_role = "id") %>%
  step_rm(any_of(c(
    "taxonname_clean",
    "taxorder", "taxsuborder",
    "taxsubgrp_mod1", "taxsubgrp_mod2", "taxsubgrp_mod3",
    "taxpartsize"
  ))) %>%
  step_zv(all_predictors()) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_normalize(all_numeric_predictors())

prep_rec_gg_inner <- prep(rec_gg_inner)

x_train_gg_inner <- bake(prep_rec_gg_inner, new_data = train_gg_inner) %>%
  select(-taxgrtgroup, -peiid)

x_val_gg_inner <- bake(prep_rec_gg_inner, new_data = val_gg_inner) %>%
  select(-taxgrtgroup, -peiid)

x_train_gg_inner <- x_train_gg_inner %>% mutate(across(where(is.logical), as.integer))
x_val_gg_inner   <- x_val_gg_inner   %>% mutate(across(where(is.logical), as.integer))

stopifnot(all(vapply(x_train_gg_inner, is.numeric, logical(1))))

# -----------------------------
# 2) Small screening grid
# -----------------------------
grid_umap_hdb_gg <- tidyr::crossing(
  n_components = c(10, 15),
  n_neighbors  = c(10, 30),
  min_dist     = c(0.01, 0.10),
  minPts       = c(5, 10)
)

cat("\n=== GREAT GROUP SCREEN GRID ===\n")
print(grid_umap_hdb_gg)

# -----------------------------
# 3) Screen combinations on inner validation set
# -----------------------------
screen_gg <- purrr::pmap_dfr(
  grid_umap_hdb_gg,
  function(n_components, n_neighbors, min_dist, minPts) {
    
    tryCatch({
      set.seed(1000 + n_components + n_neighbors + round(min_dist * 100) + minPts)
      
      umap_fit_inner <- uwot::umap(
        as.matrix(x_train_gg_inner),
        n_components = n_components,
        n_neighbors  = min(n_neighbors, nrow(x_train_gg_inner) - 1),
        min_dist     = min_dist,
        metric       = "euclidean",
        ret_model    = TRUE
      )
      
      z_train_inner <- as.data.frame(umap_fit_inner$embedding)
      names(z_train_inner) <- paste0("U", seq_len(ncol(z_train_inner)))
      
      z_val_inner <- as.data.frame(
        uwot::umap_transform(as.matrix(x_val_gg_inner), umap_fit_inner)
      )
      names(z_val_inner) <- names(z_train_inner)
      
      train_umap_inner <- bind_cols(
        z_train_inner,
        taxgrtgroup = train_gg_inner$taxgrtgroup
      )
      
      val_umap_inner <- bind_cols(
        z_val_inner,
        taxgrtgroup = val_gg_inner$taxgrtgroup
      )
      
      hdb_inner <- add_hdbscan_features(
        train_umap = train_umap_inner,
        test_umap  = val_umap_inner,
        outcome    = "taxgrtgroup",
        minPts     = minPts
      )
      
      rf_inner <- fit_tuned_rf_classifier(
        train_dat      = hdb_inner$train,
        test_dat       = hdb_inner$test,
        outcome        = "taxgrtgroup",
        trees          = 1500,
        metric_select  = "bal_accuracy",
        grid_levels    = 4,
        mtry_max       = min(ncol(hdb_inner$train) - 1, 12),
        min_n_max      = 30,
        seed_cv        = 200 + n_components + n_neighbors + minPts
      )
      
      pred_val <- rf_inner$pred
      
      met <- metric_set(accuracy, kap, bal_accuracy)(
        pred_val,
        truth = taxgrtgroup,
        estimate = .pred_class
      )
      
      f1 <- f_meas(
        pred_val,
        truth = taxgrtgroup,
        estimate = .pred_class,
        estimator = "macro"
      )
      
      conf <- confidence_summary_tbl(pred_val, "screen")
      
      tibble(
        n_components = n_components,
        n_neighbors  = n_neighbors,
        min_dist     = min_dist,
        minPts       = minPts,
        accuracy     = met$.estimate[met$.metric == "accuracy"],
        kap          = met$.estimate[met$.metric == "kap"],
        bal_accuracy = met$.estimate[met$.metric == "bal_accuracy"],
        f_meas_macro = f1$.estimate,
        mean_top1    = conf$mean_top1,
        pct_low_top1 = conf$pct_low_top1,
        mean_margin  = conf$mean_margin,
        pct_low_margin = conf$pct_low_margin,
        rf_mtry      = rf_inner$best$mtry,
        rf_min_n     = rf_inner$best$min_n,
        status       = "ok"
      )
      
    }, error = function(e) {
      tibble(
        n_components = n_components,
        n_neighbors  = n_neighbors,
        min_dist     = min_dist,
        minPts       = minPts,
        accuracy     = NA_real_,
        kap          = NA_real_,
        bal_accuracy = NA_real_,
        f_meas_macro = NA_real_,
        mean_top1    = NA_real_,
        pct_low_top1 = NA_real_,
        mean_margin  = NA_real_,
        pct_low_margin = NA_real_,
        rf_mtry      = NA_integer_,
        rf_min_n     = NA_integer_,
        status       = conditionMessage(e)
      )
    })
  }
)
beepr::beep("wilhelm")
cat("\n=== GREAT GROUP SCREEN RESULTS ===\n")
print(
  screen_gg %>%
    arrange(desc(bal_accuracy), desc(f_meas_macro), desc(accuracy))
)

best_screen_gg <- screen_gg %>%
  filter(status == "ok") %>%
  arrange(desc(bal_accuracy), desc(f_meas_macro), desc(accuracy)) %>%
  dplyr::slice(1)

cat("\n=== BEST SCREENED SETTINGS ===\n")
print(best_screen_gg)

# -----------------------------
# 4) Refit best combo on FULL train_gg, score on test_gg
# -----------------------------
best_n_components <- best_screen_gg$n_components[[1]]
best_n_neighbors  <- best_screen_gg$n_neighbors[[1]]
best_min_dist     <- best_screen_gg$min_dist[[1]]
best_minPts       <- best_screen_gg$minPts[[1]]

set.seed(130)

umap_fit_gg_tuned <- uwot::umap(
  as.matrix(x_train_gg),
  n_components = best_n_components,
  n_neighbors  = min(best_n_neighbors, nrow(x_train_gg) - 1),
  min_dist     = best_min_dist,
  metric       = "euclidean",
  ret_model    = TRUE
)

z_train_gg_tuned <- as.data.frame(umap_fit_gg_tuned$embedding)
names(z_train_gg_tuned) <- paste0("U", seq_len(ncol(z_train_gg_tuned)))

z_test_gg_tuned <- as.data.frame(
  uwot::umap_transform(as.matrix(x_test_gg), umap_fit_gg_tuned)
)
names(z_test_gg_tuned) <- names(z_train_gg_tuned)

train_umap_gg_tuned <- bind_cols(
  z_train_gg_tuned,
  taxgrtgroup = train_gg$taxgrtgroup
)

test_umap_gg_tuned <- bind_cols(
  z_test_gg_tuned,
  taxgrtgroup = test_gg$taxgrtgroup
)

hdb_gg_tuned <- add_hdbscan_features(
  train_umap = train_umap_gg_tuned,
  test_umap  = test_umap_gg_tuned,
  outcome    = "taxgrtgroup",
  minPts     = best_minPts
)

rf_gg_hdb_tuned <- fit_tuned_rf_classifier(
  train_dat      = hdb_gg_tuned$train,
  test_dat       = hdb_gg_tuned$test,
  outcome        = "taxgrtgroup",
  trees          = 1500,
  metric_select  = "bal_accuracy",
  grid_levels    = 5,
  mtry_max       = min(ncol(hdb_gg_tuned$train) - 1, 15),
  min_n_max      = 30,
  seed_cv        = 131
)

pred_gg_hdb_tuned <- rf_gg_hdb_tuned$pred

cat("\n=== TUNED GREAT GROUP HDBSCAN CONFUSION MATRIX ===\n")
print(conf_mat(pred_gg_hdb_tuned, truth = taxgrtgroup, estimate = .pred_class))

cat("\n=== TUNED GREAT GROUP HDBSCAN METRICS ===\n")
print(metric_set(accuracy, kap, bal_accuracy)(
  pred_gg_hdb_tuned,
  truth = taxgrtgroup,
  estimate = .pred_class
))
print(f_meas(
  pred_gg_hdb_tuned,
  truth = taxgrtgroup,
  estimate = .pred_class,
  estimator = "macro"
))

cat("\n=== GREAT GROUP MODEL COMPARISON ===\n")
print(bind_rows(
  metric_extract_tbl(pred_gg,           "taxgrtgroup", "umap_rf_baseline"),
  metric_extract_tbl(pred_gg_hdb,       "taxgrtgroup", "umap_hdb_rf_baseline"),
  metric_extract_tbl(pred_gg_hdb_tuned, "taxgrtgroup", "umap_hdb_rf_tuned"),
  metric_extract_tbl(res_gg$pred,       "taxgrtgroup", "rf_no_umap")
) %>%
  arrange(desc(bal_accuracy), desc(f_meas_macro), desc(accuracy)))
```

### XGBOOST GG

```{r}
library(xgboost)
#XGBOOST hypertune hdbscan umap for GG:
# ------------------------------------------------------------
# ============================================================
# GREAT GROUP: TUNED UMAP + HDBSCAN + XGBOOST
# ============================================================

# If needed on your machine:
# library(bonsai)

set.seed(140)

# ------------------------------------------------------------
# 1) Reuse best UMAP + HDBSCAN settings from your tuning screen
# ------------------------------------------------------------
best_n_components <- 15
best_n_neighbors  <- 10
best_min_dist     <- 0.01
best_minPts       <- 5

# rebuild tuned UMAP on full train/test
umap_fit_gg_xgb <- uwot::umap(
  as.matrix(x_train_gg),
  n_components = best_n_components,
  n_neighbors  = min(best_n_neighbors, nrow(x_train_gg) - 1),
  min_dist     = best_min_dist,
  metric       = "euclidean",
  ret_model    = TRUE
)

z_train_gg_xgb <- as.data.frame(umap_fit_gg_xgb$embedding)
names(z_train_gg_xgb) <- paste0("U", seq_len(ncol(z_train_gg_xgb)))

z_test_gg_xgb <- as.data.frame(
  uwot::umap_transform(as.matrix(x_test_gg), umap_fit_gg_xgb)
)
names(z_test_gg_xgb) <- names(z_train_gg_xgb)

train_umap_gg_xgb <- bind_cols(
  z_train_gg_xgb,
  taxgrtgroup = train_gg$taxgrtgroup
)

test_umap_gg_xgb <- bind_cols(
  z_test_gg_xgb,
  taxgrtgroup = test_gg$taxgrtgroup
)

# add tuned HDBSCAN features
hdb_gg_xgb <- add_hdbscan_features(
  train_umap = train_umap_gg_xgb,
  test_umap  = test_umap_gg_xgb,
  outcome    = "taxgrtgroup",
  minPts     = best_minPts
)

cat("\n=== GREAT GROUP HDBSCAN CLUSTER COUNTS FOR XGBOOST (TRAIN) ===\n")
print(count(hdb_gg_xgb$train, hdb_cluster, taxgrtgroup, sort = TRUE))

# ------------------------------------------------------------
# 2) XGBoost specification
# ------------------------------------------------------------
xgb_spec_gg <- boost_tree(
  trees = tune(),
  tree_depth = tune(),
  learn_rate = tune(),
  loss_reduction = tune(),
  min_n = tune(),
  sample_size = tune(),
  mtry = tune(),
  stop_iter = 25
) %>%
  set_engine("xgboost") %>%
  set_mode("classification")

xgb_wf_gg <- workflow() %>%
  add_model(xgb_spec_gg) %>%
  add_formula(taxgrtgroup ~ .)

# ------------------------------------------------------------
# 3) CV folds
# ------------------------------------------------------------
set.seed(141)
folds_gg_xgb <- vfold_cv(hdb_gg_xgb$train, v = 5, strata = taxgrtgroup)

# ------------------------------------------------------------
# 4) Tuning grid
# ------------------------------------------------------------
p_xgb <- ncol(hdb_gg_xgb$train) - 1

xgb_params_gg <- parameters(
  trees(range = c(500L, 2500L)),
  tree_depth(range = c(2L, 8L)),
  learn_rate(range = c(-3, -0.5)),         # log10 scale
  loss_reduction(range = c(-3, 1)),        # log10 scale
  min_n(range = c(2L, 20L)),
  sample_prop(range = c(0.5, 1.0)),
  finalize(mtry(), hdb_gg_xgb$train %>% select(-taxgrtgroup))
)

set.seed(142)
grid_xgb_gg <- grid_latin_hypercube(
  xgb_params_gg,
  size = 25
)

cat("\n=== XGBOOST GRID (first rows) ===\n")
print(head(grid_xgb_gg))

# ------------------------------------------------------------
# 5) Tune XGBoost
# ------------------------------------------------------------
set.seed(143)
tuned_xgb_gg <- tune_grid(
  xgb_wf_gg,
  resamples = folds_gg_xgb,
  grid = grid_xgb_gg,
  metrics = metric_set(bal_accuracy, kap, accuracy)
)

cat("\n=== XGBOOST TUNING METRICS ===\n")
print(collect_metrics(tuned_xgb_gg))

cat("\n=== XGBOOST TUNING NOTES ===\n")
print(collect_notes(tuned_xgb_gg))

best_xgb_gg <- select_best(tuned_xgb_gg, metric = "bal_accuracy")

cat("\n=== BEST XGBOOST SETTINGS ===\n")
print(best_xgb_gg)

# ------------------------------------------------------------
# 6) Final fit on full training data
# ------------------------------------------------------------
xgb_final_gg <- finalize_workflow(xgb_wf_gg, best_xgb_gg) %>%
  fit(hdb_gg_xgb$train)

pred_gg_hdb_xgb <- predict(xgb_final_gg, hdb_gg_xgb$test, type = "prob") %>%
  bind_cols(predict(xgb_final_gg, hdb_gg_xgb$test, type = "class")) %>%
  bind_cols(hdb_gg_xgb$test %>% select(taxgrtgroup))

cat("\n=== GREAT GROUP UMAP + HDBSCAN + XGBOOST CONFUSION MATRIX ===\n")
print(conf_mat(pred_gg_hdb_xgb, truth = taxgrtgroup, estimate = .pred_class))

cat("\n=== GREAT GROUP UMAP + HDBSCAN + XGBOOST METRICS ===\n")
print(metric_set(accuracy, kap, bal_accuracy)(
  pred_gg_hdb_xgb,
  truth = taxgrtgroup,
  estimate = .pred_class
))
print(f_meas(
  pred_gg_hdb_xgb,
  truth = taxgrtgroup,
  estimate = .pred_class,
  estimator = "macro"
))

cat("\n=== GREAT GROUP UMAP + HDBSCAN + XGBOOST CONFIDENCE ===\n")
print(confidence_summary_tbl(pred_gg_hdb_xgb, "umap_hdb_xgb_tuned"))

# ------------------------------------------------------------
# 7) Compare to your other great-group models
# ------------------------------------------------------------
cat("\n=== GREAT GROUP FULL MODEL COMPARISON ===\n")
print(
  bind_rows(
    metric_extract_tbl(pred_gg,               "taxgrtgroup", "umap_rf_baseline"),
    metric_extract_tbl(pred_gg_hdb,           "taxgrtgroup", "umap_hdb_rf_baseline"),
    metric_extract_tbl(pred_gg_hdb_tuned,     "taxgrtgroup", "umap_hdb_rf_tuned"),
    metric_extract_tbl(pred_gg_hdb_xgb,       "taxgrtgroup", "umap_hdb_xgb_tuned"),
    metric_extract_tbl(res_gg$pred,           "taxgrtgroup", "rf_no_umap")
  ) %>%
    arrange(desc(bal_accuracy), desc(f_meas_macro), desc(accuracy))
)

```
### XGBOOST GG plain

```{r}
# ============================================================
# GREAT GROUP: PLAIN XGBOOST (NO UMAP)
# ============================================================

# If needed on your machine:
# library(bonsai)

set.seed(150)

# ------------------------------------------------------------
# 1) Use the exact same cleaned train/test data as RF no UMAP
# ------------------------------------------------------------
train_gg_xgb <- clean_gg$train
test_gg_xgb  <- clean_gg$test

# same preprocessing style as fit_rf_direct()
rec_gg_xgb <- recipe(taxgrtgroup ~ ., data = train_gg_xgb) %>%
  update_role(peiid, new_role = "id") %>%
  step_zv(all_predictors()) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_normalize(all_numeric_predictors())

prep_gg_xgb <- prep(rec_gg_xgb, training = train_gg_xgb, retain = TRUE)

x_train_gg_xgb <- bake(prep_gg_xgb, new_data = train_gg_xgb) %>%
  select(-taxgrtgroup, -peiid)

x_test_gg_xgb <- bake(prep_gg_xgb, new_data = test_gg_xgb) %>%
  select(-taxgrtgroup, -peiid)

p_xgb <- ncol(x_train_gg_xgb)
if (p_xgb < 1) stop("No usable predictor columns remain for XGBoost.")

# ------------------------------------------------------------
# 2) XGBoost spec
# ------------------------------------------------------------
xgb_spec_gg_no_umap <- boost_tree(
  trees = tune(),
  tree_depth = tune(),
  learn_rate = tune(),
  loss_reduction = tune(),
  min_n = tune(),
  sample_size = tune(),
  mtry = tune(),
  stop_iter = 25
) %>%
  set_engine("xgboost") %>%
  set_mode("classification")

xgb_wf_gg_no_umap <- workflow() %>%
  add_model(xgb_spec_gg_no_umap) %>%
  add_recipe(rec_gg_xgb)

# ------------------------------------------------------------
# 3) CV folds
# ------------------------------------------------------------
set.seed(151)
folds_gg_xgb_no_umap <- vfold_cv(train_gg_xgb, v = 5, strata = taxgrtgroup)

# ------------------------------------------------------------
# 4) Tune over a Latin hypercube grid
# ------------------------------------------------------------
xgb_params_gg_no_umap <- parameters(
  trees(range = c(500L, 2500L)),
  tree_depth(range = c(2L, 8L)),
  learn_rate(range = c(-3, -0.5)),      # log10 scale
  loss_reduction(range = c(-3, 1)),     # log10 scale
  min_n(range = c(2L, 20L)),
  sample_prop(range = c(0.5, 1.0)),
  finalize(mtry(), x_train_gg_xgb)
)

set.seed(152)
grid_gg_xgb_no_umap <- grid_latin_hypercube(
  xgb_params_gg_no_umap,
  size = 25
)

cat("\n=== NO-UMAP XGBOOST GRID (first rows) ===\n")
print(head(grid_gg_xgb_no_umap))

# ------------------------------------------------------------
# 5) Tune XGBoost
# ------------------------------------------------------------
set.seed(153)
tuned_gg_xgb_no_umap <- tune_grid(
  xgb_wf_gg_no_umap,
  resamples = folds_gg_xgb_no_umap,
  grid = grid_gg_xgb_no_umap,
  metrics = metric_set(bal_accuracy, kap, accuracy)
)

cat("\n=== NO-UMAP XGBOOST TUNING METRICS ===\n")
print(collect_metrics(tuned_gg_xgb_no_umap))

cat("\n=== NO-UMAP XGBOOST TUNING NOTES ===\n")
print(collect_notes(tuned_gg_xgb_no_umap))

best_gg_xgb_no_umap <- select_best(tuned_gg_xgb_no_umap, metric = "bal_accuracy")

cat("\n=== BEST NO-UMAP XGBOOST SETTINGS ===\n")
print(best_gg_xgb_no_umap)

# ------------------------------------------------------------
# 6) Final fit on full training data
# ------------------------------------------------------------
xgb_final_gg_no_umap <- finalize_workflow(
  xgb_wf_gg_no_umap,
  best_gg_xgb_no_umap
) %>%
  fit(train_gg_xgb)

pred_gg_xgb_no_umap <- predict(xgb_final_gg_no_umap, test_gg_xgb, type = "prob") %>%
  bind_cols(predict(xgb_final_gg_no_umap, test_gg_xgb, type = "class")) %>%
  bind_cols(test_gg_xgb %>% select(taxgrtgroup))

cat("\n=== GREAT GROUP PLAIN XGBOOST (NO UMAP) CONFUSION MATRIX ===\n")
print(conf_mat(pred_gg_xgb_no_umap, truth = taxgrtgroup, estimate = .pred_class))

cat("\n=== GREAT GROUP PLAIN XGBOOST (NO UMAP) METRICS ===\n")
print(metric_set(accuracy, kap, bal_accuracy)(
  pred_gg_xgb_no_umap,
  truth = taxgrtgroup,
  estimate = .pred_class
))
print(f_meas(
  pred_gg_xgb_no_umap,
  truth = taxgrtgroup,
  estimate = .pred_class,
  estimator = "macro"
))

cat("\n=== GREAT GROUP PLAIN XGBOOST (NO UMAP) CONFIDENCE ===\n")
print(confidence_summary_tbl(pred_gg_xgb_no_umap, "xgb_no_umap"))

# ------------------------------------------------------------
# 7) Compare against your existing great-group models
# ------------------------------------------------------------
cat("\n=== GREAT GROUP: SAME TEST ROWS FOR RF NO UMAP VS XGB NO UMAP? ===\n")
print(same_truth_check(res_gg$pred, pred_gg_xgb_no_umap, "taxgrtgroup"))

cat("\n=== GREAT GROUP FULL MODEL COMPARISON (+ XGB NO UMAP) ===\n")
print(
  bind_rows(
    metric_extract_tbl(res_gg$pred,           "taxgrtgroup", "rf_no_umap"),
    metric_extract_tbl(pred_gg_xgb_no_umap,   "taxgrtgroup", "xgb_no_umap"),
    metric_extract_tbl(pred_gg_hdb_tuned,     "taxgrtgroup", "umap_hdb_rf_tuned"),
    metric_extract_tbl(pred_gg_hdb,           "taxgrtgroup", "umap_hdb_rf_baseline"),
    metric_extract_tbl(pred_gg,               "taxgrtgroup", "umap_rf_baseline"),
    metric_extract_tbl(pred_gg_hdb_xgb,       "taxgrtgroup", "umap_hdb_xgb_tuned")
  ) %>%
    arrange(desc(bal_accuracy), desc(f_meas_macro), desc(accuracy))
)
```
```{r}
#top predictors?
# ============================================================
# GREAT GROUP: TOP PREDICTORS FROM PLAIN XGBOOST (NO UMAP)
# ============================================================

library(xgboost)

# pull out fitted xgboost booster from workflow
xgb_fit_obj <- xgb_final_gg_no_umap %>%
  extract_fit_parsnip() %>%
  pluck("fit")

# feature names after recipe baking
feature_names_gg <- bake(prep_gg_xgb, new_data = train_gg_xgb) %>%
  select(-taxgrtgroup, -peiid) %>%
  names()

# xgboost importance
imp_gg_xgb <- xgboost::xgb.importance(
  feature_names = feature_names_gg,
  model = xgb_fit_obj
)

cat("\n=== TOP 20 XGBOOST PREDICTORS (GAIN) ===\n")
print(imp_gg_xgb %>% as_tibble() %>% slice_max(Gain, n = 20))



imp_gg_xgb %>%
  as_tibble() %>%
  slice_max(Gain, n = 20) %>%
  mutate(Feature = forcats::fct_reorder(Feature, Gain)) %>%
  ggplot(aes(x = Gain, y = Feature)) +
  geom_col() +
  labs(
    title = "Top 20 predictors for great-group XGBoost model",
    x = "Gain",
    y = NULL
  ) +
  theme_minimal()

library(vip)

set.seed(160)


pred_fun_multiclass <- function(object, newdata) {
  predict(object, new_data = newdata, type = "class")$.pred_class
}

vi_perm_gg <- vip::vi_permute(
  object = xgb_final_gg_no_umap,
  feature_names = names(train_gg_xgb %>% select(-taxgrtgroup)),
  train = train_gg_xgb %>% select(-taxgrtgroup),
  target = train_gg_xgb$taxgrtgroup,
  metric = "accuracy",
  pred_wrapper = pred_fun_multiclass,
  nsim = 10
)

vip::vip(vi_perm_gg, num_features = 20)

# The top predictors in the no-UMAP XGBoost great-group model were dominated by profile texture and subsurface horizon architecture. Variables describing silt, clay, and sand content, together with the presence, thickness, and depth of Bt/argillic horizons, contributed most strongly to model gain. Additional importance was associated with overall horizon composition (e.g., C, E, Bw, and Btg proportions) and wetness-related indicators such as gley features, suggesting that great-group prediction was driven primarily by whole-profile morphological organization rather than by any single chemical attribute.
```


### Taxsubgrp1:

```{r}
# ============================================================
# MODIFIER-1 MODEL (taxsubgrp_mod1) - with rare modifiers dropped
# ============================================================

set.seed(21)

min_n_mod1 <- 8

df_sg1 <- pedon_pack_150_p %>%
  mutate(
    peiid = as.character(peiid),
    taxsubgrp_mod1 = as.character(taxsubgrp_mod1),
    taxsubgrp_mod1 = str_squish(taxsubgrp_mod1),
    taxsubgrp_mod1 = na_if(taxsubgrp_mod1, "")
  ) %>%
  filter(!is.na(taxsubgrp_mod1)) %>%
  add_count(taxsubgrp_mod1, name = "class_n") %>%
  filter(class_n >= min_n_mod1) %>%
  select(-class_n) %>%
  mutate(taxsubgrp_mod1 = factor(taxsubgrp_mod1)) %>%
  droplevels()

cat("\n=== MOD1 COUNTS AFTER FILTERING ===\n")
print(count(df_sg1, taxsubgrp_mod1, sort = TRUE))

split_sg1 <- initial_split(df_sg1, prop = 0.8, strata = taxsubgrp_mod1)
train_sg1 <- training(split_sg1)
test_sg1  <- testing(split_sg1)

# drop obvious taxonomy leaks
train_sg1 <- train_sg1 %>%
  select(-any_of(c(
    "taxonname_clean", "taxonname_key", "taxclname",
    "taxorder", "taxsuborder", "taxgrtgroup", "taxpartsize",
    "taxsubgrp_mod2", "taxsubgrp_mod3"
  )))

test_sg1 <- test_sg1 %>%
  select(-any_of(c(
    "taxonname_clean", "taxonname_key", "taxclname",
    "taxorder", "taxsuborder", "taxgrtgroup", "taxpartsize",
    "taxsubgrp_mod2", "taxsubgrp_mod3"
  )))

# logical -> integer
train_sg1 <- train_sg1 %>% mutate(across(where(is.logical), as.integer))
test_sg1  <- test_sg1  %>% mutate(across(where(is.logical), as.integer))

# remove leftover character predictors except peiid
char_preds <- names(train_sg1)[vapply(train_sg1, is.character, logical(1))]
char_preds <- setdiff(char_preds, "peiid")

if (length(char_preds) > 0) {
  train_sg1 <- train_sg1 %>% select(-all_of(char_preds))
  test_sg1  <- test_sg1  %>% select(-all_of(char_preds))
}

# remove one-value predictors based on training data
pred_names <- setdiff(names(train_sg1), c("peiid", "taxsubgrp_mod1"))
one_val <- pred_names[
  vapply(train_sg1[pred_names], function(x) {
    length(unique(x[!is.na(x)])) <= 1
  }, logical(1))
]

if (length(one_val) > 0) {
  train_sg1 <- train_sg1 %>% select(-all_of(one_val))
  test_sg1  <- test_sg1  %>% select(-all_of(one_val))
}

rec_sg1 <- recipe(taxsubgrp_mod1 ~ ., data = train_sg1) %>%
  update_role(peiid, new_role = "id") %>%
  step_zv(all_predictors()) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_normalize(all_numeric_predictors())

prep_rec_sg1 <- prep(rec_sg1)

x_train_sg1 <- bake(prep_rec_sg1, new_data = train_sg1) %>%
  select(-taxsubgrp_mod1, -peiid)

x_test_sg1 <- bake(prep_rec_sg1, new_data = test_sg1) %>%
  select(-taxsubgrp_mod1, -peiid)

cat("x_train rows:", nrow(x_train_sg1), "\n")
cat("x_train cols:", ncol(x_train_sg1), "\n")

# UMAP
set.seed(22)
n_neighbors_sg1  <- min(15, nrow(x_train_sg1) - 1)
n_components_sg1 <- min(10, ncol(x_train_sg1))

umap_fit_sg1 <- uwot::umap(
  as.matrix(x_train_sg1),
  n_components = n_components_sg1,
  n_neighbors = n_neighbors_sg1,
  min_dist = 0.1,
  metric = "euclidean",
  ret_model = TRUE
)

z_train_sg1 <- as.data.frame(umap_fit_sg1$embedding)
names(z_train_sg1) <- paste0("U", seq_len(ncol(z_train_sg1)))

z_test_sg1 <- as.data.frame(uwot::umap_transform(as.matrix(x_test_sg1), umap_fit_sg1))
names(z_test_sg1) <- names(z_train_sg1)

train_umap_sg1 <- bind_cols(z_train_sg1, taxsubgrp_mod1 = train_sg1$taxsubgrp_mod1)
test_umap_sg1  <- bind_cols(z_test_sg1,  taxsubgrp_mod1 = test_sg1$taxsubgrp_mod1)

# safer CV size
min_class_train <- train_umap_sg1 %>%
  count(taxsubgrp_mod1) %>%
  summarise(min_n = min(n)) %>%
  pull(min_n)

v_sg1 <- min(5, min_class_train)
if (v_sg1 < 2) stop("At least one class still has <2 training samples after filtering.")

set.seed(23)
folds_sg1 <- vfold_cv(train_umap_sg1, v = v_sg1, strata = taxsubgrp_mod1)

rf_spec_tune_sg1 <- rand_forest(
  trees = 1500,
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("ranger", probability = TRUE, importance = "permutation") %>%
  set_mode("classification")

wf_tune_sg1 <- workflow() %>%
  add_model(rf_spec_tune_sg1) %>%
  add_formula(taxsubgrp_mod1 ~ .)

grid_sg1 <- grid_regular(
  mtry(range = c(2, min(10, ncol(z_train_sg1)))),
  min_n(range = c(2, 20)),
  levels = 4
)

tuned_sg1 <- tune_grid(
  wf_tune_sg1,
  resamples = folds_sg1,
  grid = grid_sg1,
  metrics = metric_set(accuracy, kap, bal_accuracy)
)

collect_metrics(tuned_sg1)
collect_notes(tuned_sg1)

best_sg1 <- select_best(tuned_sg1, metric = "bal_accuracy")

rf_final_sg1 <- finalize_workflow(wf_tune_sg1, best_sg1) %>%
  fit(train_umap_sg1)

pred_sg1 <- predict(rf_final_sg1, test_umap_sg1, type = "prob") %>%
  bind_cols(predict(rf_final_sg1, test_umap_sg1, type = "class")) %>%
  bind_cols(test_umap_sg1 %>% select(taxsubgrp_mod1))

cat("\n=== MOD1 METRICS (ALL ORDERS) ===\n")
print(metric_set(accuracy, kap, bal_accuracy)(
  pred_sg1,
  truth = taxsubgrp_mod1,
  estimate = .pred_class
))
print(f_meas(
  pred_sg1,
  truth = taxsubgrp_mod1,
  estimate = .pred_class,
  estimator = "macro"
))

cat("\n=== MOD1 CONFUSION MATRIX ===\n")
print(conf_mat(pred_sg1, truth = taxsubgrp_mod1, estimate = .pred_class))

cat("\n=== MOD1 CONFUSION MATRIX SUMMARY ===\n")
print(summary(conf_mat(pred_sg1, truth = taxsubgrp_mod1, estimate = .pred_class)))

cat("\n=== MOD1 TRUTH VS PREDICTED COUNTS ===\n")
print(
  pred_sg1 %>%
    count(taxsubgrp_mod1, .pred_class, sort = TRUE)
)
metric_set(accuracy, kap, bal_accuracy)(
  pred_sg1,
  truth = taxsubgrp_mod1,
  estimate = .pred_class
)

# Confidence summary
pred_conf_sg1 <- pred_sg1 %>%
 rowwise() %>%
  mutate(
    probs = list(c_across(starts_with(".pred_") & !matches("^\\.pred_class$"))),
    top1   = max(probs),
    top2   = ifelse(length(probs) > 1, sort(probs, decreasing = TRUE)[2], NA_real_),
    margin = ifelse(length(probs) > 1, top1 - top2, NA_real_)
  ) %>%
  ungroup()

cat("\n=== ULTISOLS ONLY: MOD1 CONFIDENCE ===\n")
print(
  pred_conf_sg1 %>%
    summarise(
      mean_top1      = mean(top1, na.rm = TRUE),
      pct_low_top1   = mean(top1 < 0.5, na.rm = TRUE),
      mean_margin    = mean(margin, na.rm = TRUE),
      pct_low_margin = mean(margin < 0.15, na.rm = TRUE)
    )
)

# ------------------------------------------------------------
# MOD1 (ALL ORDERS): HDBSCAN + RF
# ------------------------------------------------------------
hdb_sg1 <- add_hdbscan_features(
  train_umap = train_umap_sg1,
  test_umap  = test_umap_sg1,
  outcome    = "taxsubgrp_mod1",
  minPts     = 10
)

cat("\n=== MOD1 HDBSCAN CLUSTER COUNTS (TRAIN) ===\n")
print(count(hdb_sg1$train, hdb_cluster, taxsubgrp_mod1, sort = TRUE))

rf_sg1_hdb <- fit_tuned_rf_classifier(
  train_dat      = hdb_sg1$train,
  test_dat       = hdb_sg1$test,
  outcome        = "taxsubgrp_mod1",
  trees          = 1500,
  metric_select  = "bal_accuracy",
  grid_levels    = 4,
  mtry_max       = min(10, ncol(hdb_sg1$train) - 1),
  min_n_max      = 20,
  seed_cv        = 223
)

pred_sg1_hdb <- rf_sg1_hdb$pred

cat("\n=== MOD1 (ALL ORDERS): BASELINE VS HDBSCAN ===\n")
print(metric_compare_tbl(pred_sg1, pred_sg1_hdb, outcome = "taxsubgrp_mod1"))

cat("\n=== MOD1 (ALL ORDERS) CONFIDENCE: BASELINE VS HDBSCAN ===\n")
print(bind_rows(
  confidence_summary_tbl(pred_sg1,     "baseline"),
  confidence_summary_tbl(pred_sg1_hdb, "hdbscan")
))

cat("\n=== MOD1 (ALL ORDERS) HDBSCAN CONFUSION MATRIX ===\n")
print(conf_mat(pred_sg1_hdb, truth = taxsubgrp_mod1, estimate = .pred_class))
```



### Taxsubgrp1: Ultisols only


```{r}
# ============================================================
# ULTISOLS ONLY: taxsubgrp_mod1 prediction
# ============================================================

set.seed(31)

min_n_mod1_u <- 8   # adjust up/down if needed

df_u <- pedon_pack_150_p %>%
  mutate(
    peiid = as.character(peiid),
    taxorder = str_squish(as.character(taxorder)),
    taxsubgrp_mod1 = str_squish(as.character(taxsubgrp_mod1)),
    taxsubgrp_mod1 = na_if(taxsubgrp_mod1, "")
  ) %>%
  filter(taxorder == "Ultisols") %>%
  filter(!is.na(taxsubgrp_mod1)) %>%
  add_count(taxsubgrp_mod1, name = "class_n") %>%
  filter(class_n >= min_n_mod1_u) %>%
  select(-class_n) %>%
  mutate(taxsubgrp_mod1 = factor(taxsubgrp_mod1)) %>%
  droplevels()

cat("\n=== ULTISOLS ONLY: MOD1 COUNTS ===\n")
print(count(df_u, taxsubgrp_mod1, sort = TRUE))

split_u <- initial_split(df_u, prop = 0.8, strata = taxsubgrp_mod1)
train_u <- training(split_u)
test_u  <- testing(split_u)

# ------------------------------------------------------------
# Drop taxonomy leaks
# ------------------------------------------------------------
train_u <- train_u %>%
  select(-any_of(c(
    "taxonname_clean", "taxonname_key", "taxclname",
    "taxorder", "taxsuborder", "taxgrtgroup", "taxpartsize",
    "taxsubgrp_mod2", "taxsubgrp_mod3"
  )))

test_u <- test_u %>%
  select(-any_of(c(
    "taxonname_clean", "taxonname_key", "taxclname",
    "taxorder", "taxsuborder", "taxgrtgroup", "taxpartsize",
    "taxsubgrp_mod2", "taxsubgrp_mod3"
  )))

# logical -> integer
train_u <- train_u %>% mutate(across(where(is.logical), as.integer))
test_u  <- test_u  %>% mutate(across(where(is.logical), as.integer))

# drop leftover character predictors except peiid
char_preds_u <- names(train_u)[vapply(train_u, is.character, logical(1))]
char_preds_u <- setdiff(char_preds_u, "peiid")

if (length(char_preds_u) > 0) {
  cat("\nDropping remaining character predictors:\n")
  print(char_preds_u)
  train_u <- train_u %>% select(-all_of(char_preds_u))
  test_u  <- test_u  %>% select(-all_of(char_preds_u))
}

# drop one-value predictors based on training set
pred_names_u <- setdiff(names(train_u), c("peiid", "taxsubgrp_mod1"))

one_val_u <- pred_names_u[
  vapply(train_u[pred_names_u], function(x) {
    length(unique(x[!is.na(x)])) <= 1
  }, logical(1))
]

if (length(one_val_u) > 0) {
  cat("\nDropping one-value predictors:\n")
  print(one_val_u)
  train_u <- train_u %>% select(-all_of(one_val_u))
  test_u  <- test_u  %>% select(-all_of(one_val_u))
}

# ------------------------------------------------------------
# Recipe
# ------------------------------------------------------------
rec_u <- recipe(taxsubgrp_mod1 ~ ., data = train_u) %>%
  update_role(peiid, new_role = "id") %>%
  step_zv(all_predictors()) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_normalize(all_numeric_predictors())

prep_rec_u <- prep(rec_u)

x_train_u <- bake(prep_rec_u, new_data = train_u) %>%
  select(-taxsubgrp_mod1, -peiid)

x_test_u <- bake(prep_rec_u, new_data = test_u) %>%
  select(-taxsubgrp_mod1, -peiid)

cat("\nTrain rows:", nrow(x_train_u), "\n")
cat("Train cols:", ncol(x_train_u), "\n")

if (ncol(x_train_u) < 2) stop("Too few predictor columns left for UMAP.")
if (nrow(x_train_u) < 3) stop("Too few training rows for UMAP.")

# ------------------------------------------------------------
# UMAP
# ------------------------------------------------------------
set.seed(32)

n_neighbors_u  <- min(15, nrow(x_train_u) - 1)
n_components_u <- min(10, ncol(x_train_u))

umap_fit_u <- uwot::umap(
  as.matrix(x_train_u),
  n_components = n_components_u,
  n_neighbors = n_neighbors_u,
  min_dist = 0.1,
  metric = "euclidean",
  ret_model = TRUE
)

z_train_u <- as.data.frame(umap_fit_u$embedding)
names(z_train_u) <- paste0("U", seq_len(ncol(z_train_u)))

z_test_u <- as.data.frame(uwot::umap_transform(as.matrix(x_test_u), umap_fit_u))
names(z_test_u) <- names(z_train_u)

train_umap_u <- bind_cols(z_train_u, taxsubgrp_mod1 = train_u$taxsubgrp_mod1)
test_umap_u  <- bind_cols(z_test_u,  taxsubgrp_mod1 = test_u$taxsubgrp_mod1)

# ------------------------------------------------------------
# RF tuning
# ------------------------------------------------------------
rf_spec_tune_u <- rand_forest(
  trees = 1500,
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("ranger", probability = TRUE, importance = "permutation") %>%
  set_mode("classification")

wf_tune_u <- workflow() %>%
  add_model(rf_spec_tune_u) %>%
  add_formula(taxsubgrp_mod1 ~ .)

min_class_train_u <- train_umap_u %>%
  count(taxsubgrp_mod1) %>%
  summarise(min_n = min(n)) %>%
  pull(min_n)

v_u <- min(5, min_class_train_u)
if (v_u < 2) stop("At least one Ultisol modifier class has <2 training samples after filtering.")

set.seed(33)
folds_u <- vfold_cv(train_umap_u, v = v_u, strata = taxsubgrp_mod1)

grid_u <- grid_regular(
  mtry(range = c(2, min(10, ncol(z_train_u)))),
  min_n(range = c(2, 20)),
  levels = 4
)

tuned_u <- tune_grid(
  wf_tune_u,
  resamples = folds_u,
  grid = grid_u,
  metrics = metric_set(accuracy, kap, bal_accuracy)
)

cat("\n=== TUNING METRICS ===\n")
print(collect_metrics(tuned_u))

cat("\n=== TUNING NOTES ===\n")
print(collect_notes(tuned_u))

best_u <- select_best(tuned_u, metric = "bal_accuracy")

rf_final_u <- finalize_workflow(wf_tune_u, best_u) %>%
  fit(train_umap_u)

# ------------------------------------------------------------
# Final predictions
# ------------------------------------------------------------
pred_u <- predict(rf_final_u, test_umap_u, type = "prob") %>%
  bind_cols(predict(rf_final_u, test_umap_u, type = "class")) %>%
  bind_cols(test_umap_u %>% select(taxsubgrp_mod1))

cat("\n=== ULTISOLS ONLY: MOD1 METRICS ===\n")
print(metric_set(accuracy, kap, bal_accuracy)(
  pred_u,
  truth = taxsubgrp_mod1,
  estimate = .pred_class
))

print(f_meas(
  pred_u,
  truth = taxsubgrp_mod1,
  estimate = .pred_class,
  estimator = "macro"
))

cat("\n=== ULTISOLS ONLY: MOD1 CONFUSION MATRIX ===\n")
print(conf_mat(pred_u, truth = taxsubgrp_mod1, estimate = .pred_class))

cat("\n=== ULTISOLS ONLY: MOD1 CONFUSION MATRIX SUMMARY ===\n")
print(summary(conf_mat(pred_u, truth = taxsubgrp_mod1, estimate = .pred_class)))

cat("\n=== ULTISOLS ONLY: TRUTH VS PREDICTED COUNTS ===\n")
print(
  pred_u %>%
    count(taxsubgrp_mod1, .pred_class, sort = TRUE)
)

# ------------------------------------------------------------
# Confidence + margin
# ------------------------------------------------------------
pred_u <- pred_u %>%
  rowwise() %>%
  mutate(
    probs = list(c_across(starts_with(".pred_") & !matches("^\\.pred_class$"))),
    top1   = max(probs),
    top2   = ifelse(length(probs) > 1, sort(probs, decreasing = TRUE)[2], NA_real_),
    margin = ifelse(length(probs) > 1, top1 - top2, NA_real_)
  ) %>%
  ungroup()

cat("\n=== ULTISOLS ONLY: MOD1 CONFIDENCE ===\n")
print(
  pred_u %>%
    summarise(
      mean_top1      = mean(top1, na.rm = TRUE),
      pct_low_top1   = mean(top1 < 0.5, na.rm = TRUE),
      mean_margin    = mean(margin, na.rm = TRUE),
      pct_low_margin = mean(margin < 0.15, na.rm = TRUE)
    )
)

# ------------------------------------------------------------
# ULTISOLS ONLY MOD1: HDBSCAN + RF
# ------------------------------------------------------------
hdb_u <- add_hdbscan_features(
  train_umap = train_umap_u,
  test_umap  = test_umap_u,
  outcome    = "taxsubgrp_mod1",
  minPts     = 10
)

cat("\n=== ULTISOLS ONLY MOD1 HDBSCAN CLUSTER COUNTS (TRAIN) ===\n")
print(count(hdb_u$train, hdb_cluster, taxsubgrp_mod1, sort = TRUE))

rf_u_hdb <- fit_tuned_rf_classifier(
  train_dat      = hdb_u$train,
  test_dat       = hdb_u$test,
  outcome        = "taxsubgrp_mod1",
  trees          = 1500,
  metric_select  = "bal_accuracy",
  grid_levels    = 4,
  mtry_max       = min(10, ncol(hdb_u$train) - 1),
  min_n_max      = 20,
  seed_cv        = 323
)

pred_u_hdb <- rf_u_hdb$pred

cat("\n=== ULTISOLS ONLY MOD1: BASELINE VS HDBSCAN ===\n")
print(metric_compare_tbl(pred_u, pred_u_hdb, outcome = "taxsubgrp_mod1"))

cat("\n=== ULTISOLS ONLY MOD1 CONFIDENCE: BASELINE VS HDBSCAN ===\n")
print(bind_rows(
  confidence_summary_tbl(pred_u,     "baseline"),
  confidence_summary_tbl(pred_u_hdb, "hdbscan")
))

cat("\n=== ULTISOLS ONLY MOD1 HDBSCAN CONFUSION MATRIX ===\n")
print(conf_mat(pred_u_hdb, truth = taxsubgrp_mod1, estimate = .pred_class))
```

### Taxpartsize:


```{r}
# ============================================================
# ALL ORDERS: taxpartsize prediction
# ============================================================

set.seed(41)

min_n_ps <- 8   # raise if tuning fails; lower if too many classes get dropped

df_ps <- pedon_pack_150_p %>%
  mutate(
    peiid = as.character(peiid),
    taxpartsize = str_squish(as.character(taxpartsize)),
    taxpartsize = na_if(taxpartsize, "")
  ) %>%
  filter(!is.na(taxpartsize)) %>%
  add_count(taxpartsize, name = "class_n") %>%
  filter(class_n >= min_n_ps) %>%
  select(-class_n) %>%
  mutate(taxpartsize = factor(taxpartsize)) %>%
  droplevels()

cat("\n=== TAXPARTSIZE COUNTS (ALL ORDERS) ===\n")
print(count(df_ps, taxpartsize, sort = TRUE))

split_ps <- initial_split(df_ps, prop = 0.8, strata = taxpartsize)
train_ps <- training(split_ps)
test_ps  <- testing(split_ps)

# ------------------------------------------------------------
# Drop taxonomy leaks
# ------------------------------------------------------------
train_ps <- train_ps %>%
  select(-any_of(c(
    "taxonname_clean", "taxonname_key", "taxclname",
    "taxorder", "taxsuborder", "taxgrtgroup",
    "taxsubgrp_mod1", "taxsubgrp_mod2", "taxsubgrp_mod3"
  )))

test_ps <- test_ps %>%
  select(-any_of(c(
    "taxonname_clean", "taxonname_key", "taxclname",
    "taxorder", "taxsuborder", "taxgrtgroup",
    "taxsubgrp_mod1", "taxsubgrp_mod2", "taxsubgrp_mod3"
  )))

# logical -> integer
train_ps <- train_ps %>% mutate(across(where(is.logical), as.integer))
test_ps  <- test_ps  %>% mutate(across(where(is.logical), as.integer))

# drop leftover character predictors except peiid
char_preds_ps <- names(train_ps)[vapply(train_ps, is.character, logical(1))]
char_preds_ps <- setdiff(char_preds_ps, "peiid")

if (length(char_preds_ps) > 0) {
  cat("\nDropping remaining character predictors:\n")
  print(char_preds_ps)
  train_ps <- train_ps %>% select(-all_of(char_preds_ps))
  test_ps  <- test_ps  %>% select(-all_of(char_preds_ps))
}

# drop one-value predictors based on training set
pred_names_ps <- setdiff(names(train_ps), c("peiid", "taxpartsize"))

one_val_ps <- pred_names_ps[
  vapply(train_ps[pred_names_ps], function(x) {
    length(unique(x[!is.na(x)])) <= 1
  }, logical(1))
]

if (length(one_val_ps) > 0) {
  cat("\nDropping one-value predictors:\n")
  print(one_val_ps)
  train_ps <- train_ps %>% select(-all_of(one_val_ps))
  test_ps  <- test_ps  %>% select(-all_of(one_val_ps))
}

# ------------------------------------------------------------
# Recipe
# ------------------------------------------------------------
rec_ps <- recipe(taxpartsize ~ ., data = train_ps) %>%
  update_role(peiid, new_role = "id") %>%
  step_zv(all_predictors()) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_normalize(all_numeric_predictors())

prep_rec_ps <- prep(rec_ps)

x_train_ps <- bake(prep_rec_ps, new_data = train_ps) %>%
  select(-taxpartsize, -peiid)

x_test_ps <- bake(prep_rec_ps, new_data = test_ps) %>%
  select(-taxpartsize, -peiid)

cat("\nTrain rows:", nrow(x_train_ps), "\n")
cat("Train cols:", ncol(x_train_ps), "\n")

if (ncol(x_train_ps) < 2) stop("Too few predictor columns left for UMAP.")
if (nrow(x_train_ps) < 3) stop("Too few training rows for UMAP.")

# ------------------------------------------------------------
# UMAP
# ------------------------------------------------------------
set.seed(42)

n_neighbors_ps  <- min(15, nrow(x_train_ps) - 1)
n_components_ps <- min(10, ncol(x_train_ps))

umap_fit_ps <- uwot::umap(
  as.matrix(x_train_ps),
  n_components = n_components_ps,
  n_neighbors = n_neighbors_ps,
  min_dist = 0.1,
  metric = "euclidean",
  ret_model = TRUE
)

z_train_ps <- as.data.frame(umap_fit_ps$embedding)
names(z_train_ps) <- paste0("U", seq_len(ncol(z_train_ps)))

z_test_ps <- as.data.frame(uwot::umap_transform(as.matrix(x_test_ps), umap_fit_ps))
names(z_test_ps) <- names(z_train_ps)

train_umap_ps <- bind_cols(z_train_ps, taxpartsize = train_ps$taxpartsize)
test_umap_ps  <- bind_cols(z_test_ps,  taxpartsize = test_ps$taxpartsize)

# ------------------------------------------------------------
# RF tuning
# ------------------------------------------------------------
rf_spec_tune_ps <- rand_forest(
  trees = 1500,
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("ranger", probability = TRUE, importance = "permutation") %>%
  set_mode("classification")

wf_tune_ps <- workflow() %>%
  add_model(rf_spec_tune_ps) %>%
  add_formula(taxpartsize ~ .)

min_class_train_ps <- train_umap_ps %>%
  count(taxpartsize) %>%
  summarise(min_n = min(n)) %>%
  pull(min_n)

v_ps <- min(5, min_class_train_ps)
if (v_ps < 2) stop("At least one taxpartsize class has <2 training samples after filtering.")

set.seed(43)
folds_ps <- vfold_cv(train_umap_ps, v = v_ps, strata = taxpartsize)

grid_ps <- grid_regular(
  mtry(range = c(2, min(10, ncol(z_train_ps)))),
  min_n(range = c(2, 20)),
  levels = 4
)

tuned_ps <- tune_grid(
  wf_tune_ps,
  resamples = folds_ps,
  grid = grid_ps,
  metrics = metric_set(accuracy, kap, bal_accuracy)
)

cat("\n=== TUNING METRICS ===\n")
print(collect_metrics(tuned_ps))

cat("\n=== TUNING NOTES ===\n")
print(collect_notes(tuned_ps))

best_ps <- select_best(tuned_ps, metric = "bal_accuracy")

rf_final_ps <- finalize_workflow(wf_tune_ps, best_ps) %>%
  fit(train_umap_ps)

# ------------------------------------------------------------
# Final predictions
# ------------------------------------------------------------
pred_ps <- predict(rf_final_ps, test_umap_ps, type = "prob") %>%
  bind_cols(predict(rf_final_ps, test_umap_ps, type = "class")) %>%
  bind_cols(test_umap_ps %>% select(taxpartsize))

cat("\n=== TAXPARTSIZE METRICS (ALL ORDERS) ===\n")
print(metric_set(accuracy, kap, bal_accuracy)(
  pred_ps,
  truth = taxpartsize,
  estimate = .pred_class
))

print(f_meas(
  pred_ps,
  truth = taxpartsize,
  estimate = .pred_class,
  estimator = "macro"
))

cat("\n=== TAXPARTSIZE CONFUSION MATRIX ===\n")
print(conf_mat(pred_ps, truth = taxpartsize, estimate = .pred_class))

cat("\n=== TAXPARTSIZE CONFUSION MATRIX SUMMARY ===\n")
print(summary(conf_mat(pred_ps, truth = taxpartsize, estimate = .pred_class)))

cat("\n=== TAXPARTSIZE TRUTH VS PREDICTED COUNTS ===\n")
print(
  pred_ps %>%
    count(taxpartsize, .pred_class, sort = TRUE)
)

# ------------------------------------------------------------
# Confidence + margin
# ------------------------------------------------------------
pred_ps <- pred_ps %>%
  rowwise() %>%
  mutate(
    probs = list(c_across(starts_with(".pred_") & !matches("^\\.pred_class$"))),
    top1   = max(probs),
    top2   = ifelse(length(probs) > 1, sort(probs, decreasing = TRUE)[2], NA_real_),
    margin = ifelse(length(probs) > 1, top1 - top2, NA_real_)
  ) %>%
  ungroup()

cat("\n=== TAXPARTSIZE CONFIDENCE ===\n")
print(
  pred_ps %>%
    summarise(
      mean_top1      = mean(top1, na.rm = TRUE),
      pct_low_top1   = mean(top1 < 0.5, na.rm = TRUE),
      mean_margin    = mean(margin, na.rm = TRUE),
      pct_low_margin = mean(margin < 0.15, na.rm = TRUE)
    )
)

# ------------------------------------------------------------
# TAXPARTSIZE: HDBSCAN + RF
# ------------------------------------------------------------
hdb_ps <- add_hdbscan_features(
  train_umap = train_umap_ps,
  test_umap  = test_umap_ps,
  outcome    = "taxpartsize",
  minPts     = 10
)

cat("\n=== TAXPARTSIZE HDBSCAN CLUSTER COUNTS (TRAIN) ===\n")
print(count(hdb_ps$train, hdb_cluster, taxpartsize, sort = TRUE))

rf_ps_hdb <- fit_tuned_rf_classifier(
  train_dat      = hdb_ps$train,
  test_dat       = hdb_ps$test,
  outcome        = "taxpartsize",
  trees          = 1500,
  metric_select  = "bal_accuracy",
  grid_levels    = 4,
  mtry_max       = min(10, ncol(hdb_ps$train) - 1),
  min_n_max      = 20,
  seed_cv        = 423
)

pred_ps_hdb <- rf_ps_hdb$pred

cat("\n=== TAXPARTSIZE: BASELINE VS HDBSCAN ===\n")
print(metric_compare_tbl(pred_ps, pred_ps_hdb, outcome = "taxpartsize"))

cat("\n=== TAXPARTSIZE CONFIDENCE: BASELINE VS HDBSCAN ===\n")
print(bind_rows(
  confidence_summary_tbl(pred_ps,     "baseline"),
  confidence_summary_tbl(pred_ps_hdb, "hdbscan")
))

cat("\n=== TAXPARTSIZE HDBSCAN CONFUSION MATRIX ===\n")
print(conf_mat(pred_ps_hdb, truth = taxpartsize, estimate = .pred_class))
```
### NO UMAPS:
```{r}
# ============================================================
# SETUP + HELPERS
# ============================================================

library(tidyverse)
library(tidymodels)
library(recipes)
library(yardstick)
library(ranger)
library(beepr)

pedon_pack_150_p <- read.csv("../outputs/pedon_pack_150_02_26_26.csv")

drop_bad_predictors <- function(train_df, test_df, outcome, id_col = "peiid", drop_cols = NULL) {
  train_df <- train_df %>% select(-any_of(drop_cols))
  test_df  <- test_df  %>% select(-any_of(drop_cols))
  
  # logical -> integer
  train_df <- train_df %>% mutate(across(where(is.logical), as.integer))
  test_df  <- test_df  %>% mutate(across(where(is.logical), as.integer))
  
  # drop leftover character predictors except id/outcome
  char_preds <- names(train_df)[vapply(train_df, is.character, logical(1))]
  char_preds <- setdiff(char_preds, c(id_col, outcome))
  
  if (length(char_preds) > 0) {
    cat("\nDropping remaining character predictors:\n")
    print(char_preds)
    train_df <- train_df %>% select(-all_of(char_preds))
    test_df  <- test_df  %>% select(-all_of(char_preds))
  }
  
  # drop one-value predictors based on training set
  pred_names <- setdiff(names(train_df), c(id_col, outcome))
  
  one_val <- pred_names[
    vapply(train_df[pred_names], function(x) {
      length(unique(x[!is.na(x)])) <= 1
    }, logical(1))
  ]
  
  if (length(one_val) > 0) {
    cat("\nDropping one-value predictors:\n")
    print(one_val)
    train_df <- train_df %>% select(-all_of(one_val))
    test_df  <- test_df  %>% select(-all_of(one_val))
  }
  
  list(train = train_df, test = test_df)
}

fit_rf_direct <- function(train_df, test_df, outcome, id_col = "peiid",
                          trees = 1500, grid_levels = 4,
                          metric_for_best = "bal_accuracy") {
  
  outcome_sym <- rlang::sym(outcome)
  form <- as.formula(paste(outcome, "~ ."))
  
  rec <- recipe(form, data = train_df) %>%
    update_role(all_of(id_col), new_role = "id") %>%
    step_zv(all_predictors()) %>%
    step_impute_median(all_numeric_predictors()) %>%
    step_normalize(all_numeric_predictors())
  
  prep_rec <- prep(rec, training = train_df, retain = TRUE)
  
  x_train <- bake(prep_rec, new_data = train_df) %>%
    select(-all_of(c(outcome, id_col)))
  
  p <- ncol(x_train)
  if (p < 1) stop("No usable predictor columns remain.")
  
  min_class_n <- train_df %>%
    count(!!outcome_sym) %>%
    summarise(min_n = min(n)) %>%
    pull(min_n)
  
  v <- min(5, min_class_n)
  if (v < 2) stop("At least one class has <2 training samples.")
  
  rf_spec <- rand_forest(
    trees = trees,
    mtry = tune(),
    min_n = tune()
  ) %>%
    set_engine("ranger", probability = TRUE, importance = "permutation") %>%
    set_mode("classification")
  
  wf <- workflow() %>%
    add_model(rf_spec) %>%
    add_recipe(rec)
  
  folds <- vfold_cv(train_df, v = v, strata = !!outcome_sym)
  
  grid <- grid_regular(
    mtry(range = c(1, min(10, p))),
    min_n(range = c(2, 20)),
    levels = grid_levels
  )
  
  tuned <- tune_grid(
    wf,
    resamples = folds,
    grid = grid,
    metrics = metric_set(accuracy, kap, bal_accuracy)
  )
  
  notes <- collect_notes(tuned)
  metrics <- collect_metrics(tuned)
  
  cat("\n=== TUNING METRICS ===\n")
  print(metrics)
  
  cat("\n=== TUNING NOTES ===\n")
  print(notes)
  
  best <- select_best(tuned, metric = metric_for_best)
  
  final_fit <- finalize_workflow(wf, best) %>%
    fit(train_df)
  
  pred <- predict(final_fit, test_df, type = "prob") %>%
    bind_cols(predict(final_fit, test_df, type = "class")) %>%
    bind_cols(test_df %>% select(all_of(outcome)))
  
  cm <- conf_mat(pred, truth = !!outcome_sym, estimate = .pred_class)
  
  pred_conf <- pred %>%
    rowwise() %>%
    mutate(
      probs = list(c_across(starts_with(".pred_") & !matches("^\\.pred_class$"))),
      top1   = max(probs),
      top2   = ifelse(length(probs) > 1, sort(probs, decreasing = TRUE)[2], NA_real_),
      margin = ifelse(length(probs) > 1, top1 - top2, NA_real_)
    ) %>%
    ungroup()
  
  list(
    recipe = rec,
    tuned = tuned,
    best = best,
    final_fit = final_fit,
    pred = pred_conf,
    conf_mat = cm,
    metrics = metric_set(accuracy, kap, bal_accuracy)(
      pred, truth = !!outcome_sym, estimate = .pred_class
    ),
    f_meas = f_meas(
      pred, truth = !!outcome_sym, estimate = .pred_class, estimator = "macro"
    )
  )
}

# ============================================================
# COMPARISON HELPERS
# ============================================================

compare_models_tbl <- function(pred_a, pred_b, outcome,
                               model_a = "rf_no_umap",
                               model_b = "rf_umap_hdbscan") {
  outcome_sym <- rlang::sym(outcome)
  
  a_tbl <- bind_rows(
    metric_set(accuracy, kap, bal_accuracy)(
      pred_a, truth = !!outcome_sym, estimate = .pred_class
    ),
    precision(
      pred_a, truth = !!outcome_sym, estimate = .pred_class, estimator = "macro"
    ),
    recall(
      pred_a, truth = !!outcome_sym, estimate = .pred_class, estimator = "macro"
    ),
    f_meas(
      pred_a, truth = !!outcome_sym, estimate = .pred_class, estimator = "macro"
    )
  ) %>%
    mutate(model = model_a, .before = 1)
  
  b_tbl <- bind_rows(
    metric_set(accuracy, kap, bal_accuracy)(
      pred_b, truth = !!outcome_sym, estimate = .pred_class
    ),
    precision(
      pred_b, truth = !!outcome_sym, estimate = .pred_class, estimator = "macro"
    ),
    recall(
      pred_b, truth = !!outcome_sym, estimate = .pred_class, estimator = "macro"
    ),
    f_meas(
      pred_b, truth = !!outcome_sym, estimate = .pred_class, estimator = "macro"
    )
  ) %>%
    mutate(model = model_b, .before = 1)
  
  bind_rows(a_tbl, b_tbl)
}

same_truth_check <- function(pred_a, pred_b, outcome) {
  all(
    as.character(pred_a[[outcome]]) ==
      as.character(pred_b[[outcome]])
  )
}

```

```{r}
# ============================================================
# ALL ORDERS: taxorder prediction (NO UMAP)
# ============================================================

set.seed(1)

min_n_order <- 15

df_order <- pedon_pack_150_p %>%
  mutate(
    peiid = as.character(peiid),
    taxorder = factor(taxorder)
  ) %>%
  mutate(taxorder = forcats::fct_lump_min(taxorder, min = min_n_order, other_level = "Other")) %>%
  filter(taxorder != "Other") %>%
  droplevels()

cat("\n=== TAXORDER COUNTS ===\n")
print(count(df_order, taxorder, sort = TRUE))

split_order <- initial_split(df_order, prop = 0.8, strata = taxorder)
train_order <- training(split_order)
test_order  <- testing(split_order)

clean_order <- drop_bad_predictors(
  train_df = train_order,
  test_df  = test_order,
  outcome  = "taxorder",
  id_col   = "peiid",
  drop_cols = c(
    "taxonname_clean", "taxonname_key", "taxclname",
    "taxsuborder", "taxgrtgroup",
    "taxsubgrp_mod1", "taxsubgrp_mod2", "taxsubgrp_mod3",
    "taxpartsize"
  )
)

res_order <- fit_rf_direct(
  train_df = clean_order$train,
  test_df  = clean_order$test,
  outcome  = "taxorder",
  id_col   = "peiid"
)
beepr::beep("wilhelm")
cat("\n=== TAXORDER FINAL METRICS ===\n")
print(res_order$metrics)
print(res_order$f_meas)

cat("\n=== TAXORDER CONFUSION MATRIX ===\n")
print(res_order$conf_mat)

cat("\n=== TAXORDER CONFUSION MATRIX SUMMARY ===\n")
print(summary(res_order$conf_mat))

cat("\n=== TAXORDER TRUTH VS PREDICTED COUNTS ===\n")
print(res_order$pred %>% count(taxorder, .pred_class, sort = TRUE))

cat("\n=== TAXORDER: SAME TEST ROWS? ===\n")
print(same_truth_check(res_order$pred, pred_hdb, "taxorder"))

cat("\n=== TAXORDER: RF (NO UMAP) VS UMAP + HDBSCAN + RF ===\n")
print(compare_models_tbl(
  pred_a = res_order$pred,
  pred_b = pred_hdb,
  outcome = "taxorder",
  model_a = "rf_no_umap",
  model_b = "rf_umap_hdbscan"
))

cat("\n=== TAXORDER CONFIDENCE: RF (NO UMAP) VS UMAP + HDBSCAN + RF ===\n")
print(bind_rows(
  confidence_summary_tbl(res_order$pred, "rf_no_umap"),
  confidence_summary_tbl(pred_hdb,      "rf_umap_hdbscan")
))
```



```{r}
# ============================================================
# ALL ORDERS: taxgrtgroup prediction (NO UMAP)
# ============================================================

set.seed(11)

min_n_gg <- 20

df_gg <- pedon_pack_150_p %>%
  mutate(
    peiid = as.character(peiid),
    taxgrtgroup = factor(taxgrtgroup)
  ) %>%
  filter(!is.na(taxgrtgroup), taxgrtgroup != "") %>%
  mutate(taxgrtgroup = forcats::fct_lump_min(taxgrtgroup, min = min_n_gg, other_level = "Other")) %>%
  filter(taxgrtgroup != "Other") %>%
  droplevels()

cat("\n=== GREAT GROUP COUNTS ===\n")
print(count(df_gg, taxgrtgroup, sort = TRUE))

split_gg <- initial_split(df_gg, prop = 0.8, strata = taxgrtgroup)
train_gg <- training(split_gg)
test_gg  <- testing(split_gg)

clean_gg <- drop_bad_predictors(
  train_df = train_gg,
  test_df  = test_gg,
  outcome  = "taxgrtgroup",
  id_col   = "peiid",
  drop_cols = c(
    "taxonname_clean", "taxonname_key", "taxclname",
    "taxorder", "taxsuborder",
    "taxsubgrp_mod1", "taxsubgrp_mod2", "taxsubgrp_mod3",
    "taxpartsize"
  )
)

res_gg <- fit_rf_direct(
  train_df = clean_gg$train,
  test_df  = clean_gg$test,
  outcome  = "taxgrtgroup",
  id_col   = "peiid"
)
beepr::beep("wilhelm")

cat("\n=== GREAT GROUP FINAL METRICS ===\n")
print(res_gg$metrics)
print(res_gg$f_meas)

cat("\n=== GREAT GROUP CONFUSION MATRIX ===\n")
print(res_gg$conf_mat)

cat("\n=== GREAT GROUP CONFUSION MATRIX SUMMARY ===\n")
print(summary(res_gg$conf_mat))

cat("\n=== GREAT GROUP TRUTH VS PREDICTED COUNTS ===\n")
print(res_gg$pred %>% count(taxgrtgroup, .pred_class, sort = TRUE))

cat("\n=== GREAT GROUP: SAME TEST ROWS? ===\n")
print(same_truth_check(res_gg$pred, pred_gg_hdb, "taxgrtgroup"))

cat("\n=== GREAT GROUP: RF (NO UMAP) VS UMAP + HDBSCAN + RF ===\n")
print(compare_models_tbl(
  pred_a = res_gg$pred,
  pred_b = pred_gg_hdb,
  outcome = "taxgrtgroup",
  model_a = "rf_no_umap",
  model_b = "rf_umap_hdbscan"
))

cat("\n=== GREAT GROUP CONFIDENCE: RF (NO UMAP) VS UMAP + HDBSCAN + RF ===\n")
print(bind_rows(
  confidence_summary_tbl(res_gg$pred,    "rf_no_umap"),
  confidence_summary_tbl(pred_gg_hdb,    "rf_umap_hdbscan")
))
```

```{r}
# ============================================================
# ALL ORDERS: taxsubgrp_mod1 prediction (NO UMAP)
# ============================================================

set.seed(21)

min_n_mod1 <- 8

df_mod1 <- pedon_pack_150_p %>%
  mutate(
    peiid = as.character(peiid),
    taxsubgrp_mod1 = str_squish(as.character(taxsubgrp_mod1)),
    taxsubgrp_mod1 = na_if(taxsubgrp_mod1, "")
  ) %>%
  filter(!is.na(taxsubgrp_mod1)) %>%
  add_count(taxsubgrp_mod1, name = "class_n") %>%
  filter(class_n >= min_n_mod1) %>%
  select(-class_n) %>%
  mutate(taxsubgrp_mod1 = factor(taxsubgrp_mod1)) %>%
  droplevels()

cat("\n=== MOD1 COUNTS (ALL ORDERS) ===\n")
print(count(df_mod1, taxsubgrp_mod1, sort = TRUE))

split_mod1 <- initial_split(df_mod1, prop = 0.8, strata = taxsubgrp_mod1)
train_mod1 <- training(split_mod1)
test_mod1  <- testing(split_mod1)

clean_mod1 <- drop_bad_predictors(
  train_df = train_mod1,
  test_df  = test_mod1,
  outcome  = "taxsubgrp_mod1",
  id_col   = "peiid",
  drop_cols = c(
    "taxonname_clean", "taxonname_key", "taxclname",
    "taxorder", "taxsuborder", "taxgrtgroup",
    "taxpartsize",
    "taxsubgrp_mod2", "taxsubgrp_mod3"
  )
)

res_mod1 <- fit_rf_direct(
  train_df = clean_mod1$train,
  test_df  = clean_mod1$test,
  outcome  = "taxsubgrp_mod1",
  id_col   = "peiid"
)
beepr::beep("wilhelm")

cat("\n=== MOD1 FINAL METRICS ===\n")
print(res_mod1$metrics)
print(res_mod1$f_meas)

cat("\n=== MOD1 CONFUSION MATRIX ===\n")
print(res_mod1$conf_mat)

cat("\n=== MOD1 CONFUSION MATRIX SUMMARY ===\n")
print(summary(res_mod1$conf_mat))

cat("\n=== MOD1 TRUTH VS PREDICTED COUNTS ===\n")
print(res_mod1$pred %>% count(taxsubgrp_mod1, .pred_class, sort = TRUE))

cat("\n=== MOD1 (ALL ORDERS): SAME TEST ROWS? ===\n")
print(same_truth_check(res_mod1$pred, pred_sg1_hdb, "taxsubgrp_mod1"))

cat("\n=== MOD1 (ALL ORDERS): RF (NO UMAP) VS UMAP + HDBSCAN + RF ===\n")
print(compare_models_tbl(
  pred_a = res_mod1$pred,
  pred_b = pred_sg1_hdb,
  outcome = "taxsubgrp_mod1",
  model_a = "rf_no_umap",
  model_b = "rf_umap_hdbscan"
))

cat("\n=== MOD1 (ALL ORDERS) CONFIDENCE: RF (NO UMAP) VS UMAP + HDBSCAN + RF ===\n")
print(bind_rows(
  confidence_summary_tbl(res_mod1$pred,  "rf_no_umap"),
  confidence_summary_tbl(pred_sg1_hdb,   "rf_umap_hdbscan")
))
```

```{r}
# ============================================================
# ULTISOLS ONLY: taxsubgrp_mod1 prediction (NO UMAP)
# ============================================================

set.seed(31)

min_n_mod1_u <- 8

df_mod1_u <- pedon_pack_150_p %>%
  mutate(
    peiid = as.character(peiid),
    taxorder = str_squish(as.character(taxorder)),
    taxsubgrp_mod1 = str_squish(as.character(taxsubgrp_mod1)),
    taxsubgrp_mod1 = na_if(taxsubgrp_mod1, "")
  ) %>%
  filter(taxorder == "Ultisols") %>%
  filter(!is.na(taxsubgrp_mod1)) %>%
  add_count(taxsubgrp_mod1, name = "class_n") %>%
  filter(class_n >= min_n_mod1_u) %>%
  select(-class_n) %>%
  mutate(taxsubgrp_mod1 = factor(taxsubgrp_mod1)) %>%
  droplevels()

cat("\n=== ULTISOLS ONLY: MOD1 COUNTS ===\n")
print(count(df_mod1_u, taxsubgrp_mod1, sort = TRUE))

split_mod1_u <- initial_split(df_mod1_u, prop = 0.8, strata = taxsubgrp_mod1)
train_mod1_u <- training(split_mod1_u)
test_mod1_u  <- testing(split_mod1_u)

clean_mod1_u <- drop_bad_predictors(
  train_df = train_mod1_u,
  test_df  = test_mod1_u,
  outcome  = "taxsubgrp_mod1",
  id_col   = "peiid",
  drop_cols = c(
    "taxonname_clean", "taxonname_key", "taxclname",
    "taxorder", "taxsuborder", "taxgrtgroup",
    "taxpartsize",
    "taxsubgrp_mod2", "taxsubgrp_mod3"
  )
)

res_mod1_u <- fit_rf_direct(
  train_df = clean_mod1_u$train,
  test_df  = clean_mod1_u$test,
  outcome  = "taxsubgrp_mod1",
  id_col   = "peiid"
)
beepr::beep("wilhelm")

cat("\n=== ULTISOLS ONLY: MOD1 FINAL METRICS ===\n")
print(res_mod1_u$metrics)
print(res_mod1_u$f_meas)

cat("\n=== ULTISOLS ONLY: MOD1 CONFUSION MATRIX ===\n")
print(res_mod1_u$conf_mat)

cat("\n=== ULTISOLS ONLY: MOD1 CONFUSION MATRIX SUMMARY ===\n")
print(summary(res_mod1_u$conf_mat))

cat("\n=== ULTISOLS ONLY: MOD1 TRUTH VS PREDICTED COUNTS ===\n")
print(res_mod1_u$pred %>% count(taxsubgrp_mod1, .pred_class, sort = TRUE))

cat("\n=== ULTISOLS ONLY MOD1: SAME TEST ROWS? ===\n")
print(same_truth_check(res_mod1_u$pred, pred_u_hdb, "taxsubgrp_mod1"))

cat("\n=== ULTISOLS ONLY MOD1: RF (NO UMAP) VS UMAP + HDBSCAN + RF ===\n")
print(compare_models_tbl(
  pred_a = res_mod1_u$pred,
  pred_b = pred_u_hdb,
  outcome = "taxsubgrp_mod1",
  model_a = "rf_no_umap",
  model_b = "rf_umap_hdbscan"
))

cat("\n=== ULTISOLS ONLY MOD1 CONFIDENCE: RF (NO UMAP) VS UMAP + HDBSCAN + RF ===\n")
print(bind_rows(
  confidence_summary_tbl(res_mod1_u$pred, "rf_no_umap"),
  confidence_summary_tbl(pred_u_hdb,      "rf_umap_hdbscan")
))
```

```{r}
# ============================================================
# ALL ORDERS: taxpartsize prediction (NO UMAP)
# ============================================================

set.seed(41)

min_n_ps <- 8

df_ps <- pedon_pack_150_p %>%
  mutate(
    peiid = as.character(peiid),
    taxpartsize = str_squish(as.character(taxpartsize)),
    taxpartsize = na_if(taxpartsize, "")
  ) %>%
  filter(!is.na(taxpartsize)) %>%
  add_count(taxpartsize, name = "class_n") %>%
  filter(class_n >= min_n_ps) %>%
  select(-class_n) %>%
  mutate(taxpartsize = factor(taxpartsize)) %>%
  droplevels()

cat("\n=== TAXPARTSIZE COUNTS ===\n")
print(count(df_ps, taxpartsize, sort = TRUE))

split_ps <- initial_split(df_ps, prop = 0.8, strata = taxpartsize)
train_ps <- training(split_ps)
test_ps  <- testing(split_ps)

clean_ps <- drop_bad_predictors(
  train_df = train_ps,
  test_df  = test_ps,
  outcome  = "taxpartsize",
  id_col   = "peiid",
  drop_cols = c(
    "taxonname_clean", "taxonname_key", "taxclname",
    "taxorder", "taxsuborder", "taxgrtgroup",
    "taxsubgrp_mod1", "taxsubgrp_mod2", "taxsubgrp_mod3"
  )
)

res_ps <- fit_rf_direct(
  train_df = clean_ps$train,
  test_df  = clean_ps$test,
  outcome  = "taxpartsize",
  id_col   = "peiid"
)
beepr::beep("wilhelm")

cat("\n=== TAXPARTSIZE FINAL METRICS ===\n")
print(res_ps$metrics)
print(res_ps$f_meas)

cat("\n=== TAXPARTSIZE CONFUSION MATRIX ===\n")
print(res_ps$conf_mat)

cat("\n=== TAXPARTSIZE CONFUSION MATRIX SUMMARY ===\n")
print(summary(res_ps$conf_mat))

cat("\n=== TAXPARTSIZE TRUTH VS PREDICTED COUNTS ===\n")
print(res_ps$pred %>% count(taxpartsize, .pred_class, sort = TRUE))

cat("\n=== TAXPARTSIZE: SAME TEST ROWS? ===\n")
print(same_truth_check(res_ps$pred, pred_ps_hdb, "taxpartsize"))

cat("\n=== TAXPARTSIZE: RF (NO UMAP) VS UMAP + HDBSCAN + RF ===\n")
print(compare_models_tbl(
  pred_a = res_ps$pred,
  pred_b = pred_ps_hdb,
  outcome = "taxpartsize",
  model_a = "rf_no_umap",
  model_b = "rf_umap_hdbscan"
))

cat("\n=== TAXPARTSIZE CONFIDENCE: RF (NO UMAP) VS UMAP + HDBSCAN + RF ===\n")
print(bind_rows(
  confidence_summary_tbl(res_ps$pred,   "rf_no_umap"),
  confidence_summary_tbl(pred_ps_hdb,   "rf_umap_hdbscan")
))
```

```{r}
#saving models for Perry:

```



