---
title: "NCSS_spatial_series_LOOCV_RK"
format: html
---

```{r setup}
library(tidyverse)
library(terra)
library(sf)
library(gstat)
library(ranger)
library(yardstick)
library(tidyterra)
library(caret)

dir.create("../outputs/spatial", showWarnings = FALSE, recursive = TRUE)
```

```{r load-inputs}
# ============================================================
# LOAD ALL INPUTS
# ============================================================

# Covariate stack
if (!exists("cov_stack")) {
  cov_stack <- terra::rast(
    "../data/spatial/covariate_stack_3m_all_with_SR_indices.tif"
  )
}

# Pedon points
if (!exists("soil_pts")) {
  soil_pts <- sf::st_read(
    "../data/spatial/perry_pedon_locs.shp", quiet = TRUE
  ) %>%
    sf::st_transform(crs = sf::st_crs(terra::crs(cov_stack))) %>%
    rename(Point = Point_ID) %>%
    mutate(Point = as.character(Point))
}

# Covariates already extracted at pedon locations
if (!exists("cov_at_pedons")) {
  cov_at_pedons <- read.csv(
    "../outputs/spatial/covariates_at_pedons.csv"
  ) %>%
    mutate(Point = as.character(Point))
}

# Perry series truth
perry_fp_clean <- read.csv(
  "../data/Perry_FP_data_with_taxa_02_20_2026.csv"
) %>%
  mutate(Point = as.character(Point))

perry_series_truth <- perry_fp_clean %>%
  mutate(
    series = na_if(str_squish(as.character(series_clean)), ""),
    series = na_if(series, "Fill"),
    series = na_if(series, "fill"),
    series = str_to_title(series)
  ) %>%
  group_by(Point) %>%
  summarise(
    series = {
      y <- series[!is.na(series)]
      if (length(y) == 0) NA_character_
      else names(sort(table(y), decreasing = TRUE))[1]
    },
    .groups = "drop"
  )

# XY coordinates
pts_xy <- soil_pts %>%
  st_drop_geometry() %>%
  bind_cols(
    st_coordinates(soil_pts) %>%
      as.data.frame() %>%
      rename(coord_X = X, coord_Y = Y)
  ) %>%
  mutate(Point = as.character(Point)) %>%
  select(Point, coord_X, coord_Y)

cat("Inputs loaded.\n")
cat("Pedon points:", nrow(soil_pts), "\n")
cat("Covariate rows:", nrow(cov_at_pedons), "\n")
cat("Series truth rows:", nrow(perry_series_truth), "\n")
```

```{r build-model-df}
# ============================================================
# BUILD ONE-ROW-PER-PEDON MODEL DATASET
# ============================================================


# Collapse covariates to one row per Point
cov_dedup <- cov_at_pedons %>%
  mutate(Point = as.character(Point)) %>%
  group_by(Point) %>%
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)),
            .groups = "drop")

# Series truth: one row per Point
truth_dedup <- perry_series_truth %>%
  mutate(Point = as.character(Point)) %>%
  distinct(Point, .keep_all = TRUE)

# XY coordinates: extract directly from geometry to avoid bind_cols duplication
# (XYZ geometry was causing extra rows via bind_cols approach)
pts_xy_dedup <- soil_pts %>%
  st_drop_geometry() %>%
  mutate(
    coord_X = sf::st_coordinates(soil_pts)[, 1],
    coord_Y = sf::st_coordinates(soil_pts)[, 2],
    Point   = as.character(Point)
  ) %>%
  select(Point, coord_X, coord_Y) %>%
  distinct(Point, .keep_all = TRUE)

# Diagnostics
cat("cov_dedup rows:", nrow(cov_dedup), "| unique Points:", n_distinct(cov_dedup$Point), "\n")
cat("truth_dedup rows:", nrow(truth_dedup), "| unique Points:", n_distinct(truth_dedup$Point), "\n")
cat("pts_xy_dedup rows:", nrow(pts_xy_dedup), "| unique Points:", n_distinct(pts_xy_dedup$Point), "\n")

# Join everything
model_df <- cov_dedup %>%
  left_join(truth_dedup, by = "Point") %>%
  left_join(pts_xy_dedup, by = "Point") %>%
  filter(!is.na(series), !is.na(coord_X)) %>%
  add_count(series, name = "series_n") %>%
  filter(series_n >= 3) %>%
  select(-series_n) %>%
  mutate(series = factor(series))

# Hard stop if duplicates remain
stopifnot(
  "Duplicate Points remain in model_df!" =
    n_distinct(model_df$Point) == nrow(model_df)
)

cat("\nmodel_df rows:", nrow(model_df), "\n")
cat("Unique Points:", n_distinct(model_df$Point), "\n")
cat("Series distribution:\n")
model_df %>%
  mutate(series = as.character(series)) %>%
  count(series, sort = TRUE) %>%
  as.data.frame() %>%
  print()

# Covariate columns
cov_cols <- setdiff(
  names(model_df),
  c("Point", "series", "coord_X", "coord_Y")
)
cov_cols <- cov_cols[
  vapply(model_df[cov_cols],
         function(x) length(unique(x[is.finite(x)])) > 1,
         logical(1))
]
cat("Covariates:", length(cov_cols), "\n")
```

Then re-run `loocv`, `loocv-metrics`, `fit-final-rf`, and everything after in order. The `stopifnot` will catch any remaining duplication before it silently inflates accuracy again!


```{r loocv}
# ============================================================
# LOOCV: RF ON SPATIAL COVARIATES
# Train on n-1 pedons, predict the left-out pedon
# Repeat for all n pedons
# ============================================================

n <- nrow(model_df)
series_levels <- levels(model_df$series)
n_series <- length(series_levels)

# Storage for LOOCV predictions
loocv_results <- tibble(
  Point       = character(n),
  true_series = character(n),
  pred_series = character(n),
  top1_prob   = numeric(n)
)

# Storage for predicted probabilities (for kriging residuals later)
loocv_probs <- matrix(NA_real_, nrow = n, ncol = n_series,
                      dimnames = list(NULL, paste0("prob_", series_levels)))

cat("Running LOOCV (n =", n, ")...\n")

for (i in seq_len(n)) {

  train_i <- model_df[-i, ]
  test_i  <- model_df[i,  ]

  # Impute NAs with column medians from training set
  train_x <- train_i %>% select(all_of(cov_cols))
  test_x  <- test_i  %>% select(all_of(cov_cols))

  col_medians <- vapply(train_x, median, numeric(1), na.rm = TRUE)

  for (nm in cov_cols) {
    train_x[[nm]][is.na(train_x[[nm]])] <- col_medians[nm]
    test_x[[nm]][is.na(test_x[[nm]])]   <- col_medians[nm]
  }

  # Fit RF
  rf_i <- ranger::ranger(
    x          = train_x,
    y          = train_i$series,
    num.trees  = 1000,
    mtry       = max(2, floor(sqrt(length(cov_cols)))),
    min.node.size = 2,
    probability = TRUE,
    seed       = 42 + i
  )

  # Predict left-out pedon
  pred_i <- predict(rf_i, data = test_x)$predictions

  # Align probability columns to series_levels
  prob_row <- rep(NA_real_, n_series)
  names(prob_row) <- series_levels
  matched <- intersect(series_levels, colnames(pred_i))
  prob_row[matched] <- pred_i[1, matched]

  loocv_probs[i, ] <- prob_row

  pred_class <- series_levels[which.max(prob_row)]
  top1       <- max(prob_row, na.rm = TRUE)

  loocv_results$Point[i]       <- model_df$Point[i]
  loocv_results$true_series[i] <- as.character(model_df$series[i])
  loocv_results$pred_series[i] <- pred_class
  loocv_results$top1_prob[i]   <- top1

  if (i %% 10 == 0 || i == n)
    cat("  Fold", i, "/", n, "\n")
}

cat("\nLOOCV complete.\n")
```

```{r loocv-metrics}
# ============================================================
# LOOCV METRICS
# ============================================================

loocv_results <- loocv_results %>%
  mutate(
    correct = true_series == pred_series,
    true_series = factor(true_series, levels = series_levels),
    pred_series = factor(pred_series, levels = series_levels)
  )

# Overall metrics
overall_acc <- mean(loocv_results$correct, na.rm = TRUE)
overall_kap <- yardstick::kap_vec(
  truth    = loocv_results$true_series,
  estimate = loocv_results$pred_series
)
overall_f1 <- yardstick::f_meas_vec(
  truth     = loocv_results$true_series,
  estimate  = loocv_results$pred_series,
  estimator = "macro"
)

cat("\n============================================================\n")
cat("LOOCV RESULTS\n")
cat("============================================================\n")
cat("Overall accuracy:", round(overall_acc, 3), "\n")
cat("Kappa:           ", round(overall_kap, 3), "\n")
cat("Macro F1:        ", round(overall_f1,  3), "\n")

# Per-class accuracy
cat("\nPer-class accuracy:\n")
loocv_results %>%
  group_by(true_series) %>%
  summarise(
    n         = n(),
    n_correct = sum(correct),
    acc       = round(n_correct / n, 2),
    mean_prob = round(mean(top1_prob), 2),
    .groups   = "drop"
  ) %>%
  arrange(desc(acc)) %>%
  as.data.frame() %>%
  print()

# Confusion matrix
cat("\nConfusion matrix:\n")
conf <- table(
  True      = loocv_results$true_series,
  Predicted = loocv_results$pred_series
)
print(conf)

# Misclassification pairs
cat("\nMisclassification pairs:\n")
loocv_results %>%
  filter(!correct) %>%
  count(true_series, pred_series, sort = TRUE) %>%
  as.data.frame() %>%
  print()

# Save LOOCV results
write.csv(loocv_results,
          "../outputs/spatial/loocv_results.csv",
          row.names = FALSE)
```

```{r fit-final-rf}
# ============================================================
# FIT FINAL RF ON ALL PEDONS
# (used for wall-to-wall prediction + residual kriging)
# ============================================================

# Impute NAs
full_x <- model_df %>% select(all_of(cov_cols))
col_medians_full <- vapply(full_x, median, numeric(1), na.rm = TRUE)
for (nm in cov_cols) {
  full_x[[nm]][is.na(full_x[[nm]])] <- col_medians_full[nm]
}

set.seed(42)
rf_final <- ranger::ranger(
  x             = full_x,
  y             = model_df$series,
  num.trees     = 1500,
  mtry          = max(2, floor(sqrt(length(cov_cols)))),
  min.node.size = 2,
  probability   = TRUE,
  importance    = "permutation",
  seed          = 42
)

cat("Final RF fitted on all", nrow(model_df), "pedons.\n")
cat("OOB prediction error:", round(rf_final$prediction.error, 3), "\n")

# In-sample predicted probabilities (for residual kriging)
insample_probs <- rf_final$predictions
colnames(insample_probs) <- colnames(rf_final$predictions)
```

```{r variogram-kriging}
# ============================================================
# REGRESSION KRIGING: FIT VARIOGRAMS ON RF RESIDUALS
# Residual = observed indicator (0/1) - RF predicted probability
# ============================================================

variogram_fits <- list()
krige_models   <- list()

for (s in series_levels) {

  if (!s %in% colnames(insample_probs)) next

  resid_df <- tibble(
    coord_X  = model_df$coord_X,
    coord_Y  = model_df$coord_Y,
    observed = as.integer(as.character(model_df$series) == s),
    rf_prob  = insample_probs[, s],
    residual = observed - rf_prob
  ) %>%
    filter(is.finite(residual))

  if (nrow(resid_df) < 5) {
    cat("Skipping", s, "-- fewer than 5 observations.\n")
    next
  }

  resid_sf <- sf::st_as_sf(
    resid_df,
    coords = c("coord_X", "coord_Y"),
    crs    = sf::st_crs(soil_pts)
  )

  # Empirical variogram
  vgm_emp <- tryCatch(
    gstat::variogram(residual ~ 1, data = resid_sf),
    error = function(e) NULL
  )
  if (is.null(vgm_emp)) next

  # Fit theoretical variogram
  vgm_fit <- tryCatch(
    gstat::fit.variogram(vgm_emp, gstat::vgm(c("Sph", "Exp"))),
    error = function(e) tryCatch(
      gstat::fit.variogram(vgm_emp, gstat::vgm("Exp")),
      error = function(e2) NULL
    )
  )
  if (is.null(vgm_fit)) next

  variogram_fits[[s]] <- list(empirical = vgm_emp, fitted = vgm_fit)
  krige_models[[s]]   <- list(data = resid_sf, vgm = vgm_fit)

  cat("Variogram fitted:", s,
      "| Model:", as.character(vgm_fit$model[2]),
      "| Range:", round(vgm_fit$range[2], 1),
      "| Sill:", round(sum(vgm_fit$psill), 4), "\n")
}

cat("\nVariograms fitted for", length(krige_models), "series.\n")

# Plot variograms
for (s in names(variogram_fits)) {
  p <- plot(
    variogram_fits[[s]]$empirical,
    variogram_fits[[s]]$fitted,
    main = paste("Variogram of RF residuals:", s)
  )
  print(p)
}
```

```{r wall-to-wall}
# ============================================================
# WALL-TO-WALL PREDICTION
# RF probability + kriged residual correction
# ============================================================

cat("Building pixel prediction data frame...\n")
cov_df <- terra::as.data.frame(
  cov_stack[[cov_cols]], xy = TRUE, na.rm = FALSE
)
n_pixels <- nrow(cov_df)
cat("Total pixels:", n_pixels, "\n")

# Impute NAs using full-dataset medians
cov_pred <- cov_df %>%
  select(all_of(cov_cols)) %>%
  mutate(across(everything(), ~ ifelse(is.finite(.x), .x, NA_real_)))

for (nm in cov_cols) {
  cov_pred[[nm]][is.na(cov_pred[[nm]])] <- col_medians_full[nm]
}

# Predict in chunks
chunk_size <- 50000
n_chunks   <- ceiling(n_pixels / chunk_size)
prob_list  <- vector("list", n_chunks)

cat("Predicting in", n_chunks, "chunks...\n")
for (i in seq_len(n_chunks)) {
  idx <- ((i-1)*chunk_size + 1):min(i*chunk_size, n_pixels)

  chunk_preds <- predict(
    rf_final,
    data = cov_pred[idx, , drop = FALSE]
  )$predictions

  prob_list[[i]] <- as.data.frame(chunk_preds)
  if (i %% 5 == 0 || i == n_chunks)
    cat("  Chunk", i, "/", n_chunks, "\n")
}

all_probs <- bind_rows(prob_list)
cat("Prediction complete. Columns:", paste(names(all_probs), collapse=", "), "\n")

# Build RF probability raster stack
prob_raster_list <- list()

for (s in series_levels) {
  if (!s %in% names(all_probs)) next

  r <- terra::rast(cov_stack[[1]])
  terra::values(r) <- all_probs[[s]]
  names(r) <- paste0("prob_", s)

  if (any(is.finite(terra::values(r)))) {
    prob_raster_list[[s]] <- r
  }
}

prob_rasters <- terra::rast(prob_raster_list)
cat("RF probability rasters:", terra::nlyr(prob_rasters), "layers\n")
```

```{r krige-correction}
# Only krige series with enough pedons to fit a reliable variogram
# Minimum 5 pedons AND reasonable range (not > half the field extent)

# Get field extent to check range reasonableness
field_extent <- terra::ext(cov_stack)
field_diag <- sqrt((field_extent$xmax - field_extent$xmin)^2 + 
                   (field_extent$ymax - field_extent$ymin)^2)
max_reasonable_range <- field_diag * 0.5
cat("Field diagonal:", round(field_diag), "map units\n")
cat("Max reasonable variogram range:", round(max_reasonable_range), "\n")

# Filter to reliable variogram models only
reliable_krige <- names(krige_models)[
  vapply(names(krige_models), function(s) {
    n_pts <- nrow(krige_models[[s]]$data)
    range <- variogram_fits[[s]]$fitted$range[2]
    n_pts >= 5 & range < max_reasonable_range & range > 0
  }, logical(1))
]

cat("Series with reliable variograms:", paste(reliable_krige, collapse = ", "), "\n")
cat("Skipping (unreliable):", 
    paste(setdiff(names(krige_models), reliable_krige), collapse = ", "), "\n")

if (length(reliable_krige) > 0) {
  pred_grid <- terra::as.data.frame(
    cov_stack[[1]], xy = TRUE, na.rm = TRUE
  ) %>%
    select(x, y) %>%
    sf::st_as_sf(coords = c("x", "y"), crs = sf::st_crs(soil_pts))

  for (s in reliable_krige) {
    cat("Kriging:", s, "...\n")

    krige_result <- tryCatch(
      gstat::krige(
        formula   = residual ~ 1,
        locations = krige_models[[s]]$data,
        newdata   = pred_grid,
        model     = krige_models[[s]]$vgm,
        debug.level = 0
      ),
      error = function(e) {
        cat("  Failed:", conditionMessage(e), "\n")
        NULL
      }
    )
    if (is.null(krige_result)) next

    krige_sf <- krige_result %>%
      mutate(krige_val = var1.pred) %>%
      select(krige_val)

    krige_rast <- terra::rasterize(
      terra::vect(krige_sf),
      cov_stack[[1]],
      field = "krige_val",
      fun   = "mean"
    )

    layer_nm <- paste0("prob_", s)
    if (layer_nm %in% names(prob_rasters)) {
      corrected <- terra::clamp(
        prob_rasters[[layer_nm]] + krige_rast,
        lower = 0, upper = 1
      )
      prob_rasters[[layer_nm]] <- corrected
      cat("  Done.\n")
    }
  }
} else {
  cat("No reliable variograms -- using RF probabilities only.\n")
}

# Rebuild classified map from corrected probabilities
predicted_class_idx <- terra::which.lyr(prob_rasters)
names(predicted_class_idx) <- "predicted_series_idx"
fitted_series <- str_remove(names(prob_rasters), "^prob_")

# Quick sanity check
pred_vals <- terra::values(predicted_class_idx)
pred_vals_named <- fitted_series[pred_vals]
cat("\nPixel counts after correction:\n")
table(pred_vals_named, useNA = "ifany") %>%
  sort(decreasing = TRUE) %>%
  print()
```

Greenville's variogram range is 8761m which is likely larger than the field — so it'll also be filtered out. That's actually fine and honest — it means Greenville's residuals show no spatial structure at the field scale, which makes sense for a dominant series that occurs everywhere. The RF probabilities alone will drive the Greenville prediction, which is correct behavior.

Greenville's variogram range is 8761m which is likely larger than the field — so it'll also be filtered out. That's actually fine and honest — it means Greenville's residuals show no spatial structure at the field scale, which makes sense for a dominant series that occurs everywhere. The RF probabilities alone will drive the Greenville prediction, which is correct behavior.

```{r visualize}
# ============================================================
# VISUALIZATION
# ============================================================

# Factor raster for ggplot
predicted_factor <- terra::as.factor(predicted_class_idx)
levels(predicted_factor) <- data.frame(
  value  = seq_along(fitted_series),
  series = fitted_series
)

# 1. Predicted series map
p_map <- ggplot() +
  tidyterra::geom_spatraster(data = predicted_factor) +
  scale_fill_viridis_d(
    option   = "turbo",
    na.value = "grey90",
    name     = "Predicted Series"
  ) +
  theme_minimal() +
  labs(
    title    = "Predicted Soil Series — Perry Site",
    subtitle = "RF + Regression Kriging (LOOCV validated, 3m resolution)"
  )
print(p_map)
ggsave("../outputs/spatial/map_predicted_series_loocv_rk.png",
       p_map, width = 10, height = 8, dpi = 200)

# 2. Probability surface per series
for (s in fitted_series) {
  layer_nm <- paste0("prob_", s)
  if (!layer_nm %in% names(prob_rasters)) next

  p <- ggplot() +
    tidyterra::geom_spatraster(data = prob_rasters[[layer_nm]]) +
    scale_fill_distiller(
      palette   = "RdYlGn",
      direction = 1,
      na.value  = "grey90",
      limits    = c(0, 1),
      name      = "Probability"
    ) +
    theme_minimal() +
    labs(
      title    = paste("Probability surface:", s),
      subtitle = "RF + kriging correction"
    )
  print(p)
  ggsave(
    paste0("../outputs/spatial/prob_",
           str_replace_all(s, " ", "_"), "_loocv_rk.png"),
    p, width = 8, height = 6, dpi = 200
  )
}

# 3. Validation overlay: LOOCV predictions at pedon locations
loocv_sf <- loocv_results %>%
  left_join(pts_xy, by = "Point") %>%
  filter(!is.na(coord_X)) %>%
  sf::st_as_sf(coords = c("coord_X", "coord_Y"),
               crs = sf::st_crs(soil_pts))

p_val <- ggplot() +
  tidyterra::geom_spatraster(data = predicted_factor) +
  scale_fill_viridis_d(option = "turbo", na.value = "grey90",
                       name = "Predicted (map)") +
  geom_sf(data = loocv_sf,
          aes(color = true_series, shape = correct),
          size = 3, stroke = 1.2) +
  scale_color_brewer(palette = "Set1", name = "True Series") +
  scale_shape_manual(
    values = c("TRUE" = 16, "FALSE" = 4),
    name   = "LOOCV correct?"
  ) +
  theme_minimal() +
  labs(
    title    = "LOOCV Validation Overlay",
    subtitle = "Circles = correct prediction; X = misclassified"
  )
print(p_val)
ggsave("../outputs/spatial/map_loocv_validation_overlay.png",
       p_val, width = 10, height = 8, dpi = 200)

# 4. Variable importance
importance_df <- tibble(
  variable   = names(rf_final$variable.importance),
  importance = rf_final$variable.importance
) %>%
  arrange(desc(importance)) %>%
  mutate(variable = fct_reorder(variable, importance))

p_imp <- ggplot(importance_df %>% slice_head(n = 20),
                aes(x = importance, y = variable)) +
  geom_col(fill = "#2c7bb6") +
  theme_minimal() +
  labs(
    title = "Variable Importance — Spatial RF (Perry Series)",
    x     = "Permutation Importance",
    y     = NULL
  )
print(p_imp)
ggsave("../outputs/spatial/variable_importance_loocv_rf.png",
       p_imp, width = 8, height = 6, dpi = 200)
```

```{r summary}
# ============================================================
# FINAL SUMMARY
# ============================================================

cat("\n============================================================\n")
cat("SPATIAL SERIES PREDICTION — FINAL SUMMARY\n")
cat("============================================================\n")
cat("Method:            RF + Regression Kriging\n")
cat("Validation:        Leave-One-Out Cross-Validation (LOOCV)\n")
cat("Pedons:           ", nrow(model_df), "\n")
cat("Series modeled:   ", length(fitted_series), "\n")
cat("Covariates used:  ", length(cov_cols), "\n")
cat("Variograms fitted:", length(krige_models), "\n")
cat("\nLOOCV Performance:\n")
cat("  Accuracy:", round(overall_acc, 3), "\n")
cat("  Kappa:   ", round(overall_kap, 3), "\n")
cat("  Macro F1:", round(overall_f1,  3), "\n")
cat("\nOOB Error (full model):", round(rf_final$prediction.error, 3), "\n")

if (length(variogram_fits) > 0) {
  cat("\nVariogram ranges (spatial autocorrelation distance):\n")
  for (s in names(variogram_fits)) {
    cat(" ", s, ":",
        round(variogram_fits[[s]]$fitted$range[2], 1),
        "map units\n")
  }
}

cat("\nTop 10 most important covariates:\n")
importance_df %>%
  slice_head(n = 10) %>%
  mutate(importance = round(importance, 4)) %>%
  as.data.frame() %>%
  print()

# Check what the wall-to-wall map is actually predicting
cat("Pixel counts per predicted series:\n")
pred_vals <- terra::values(predicted_class_idx)
pred_vals_named <- fitted_series[pred_vals]
table(pred_vals_named, useNA = "ifany") %>%
  sort(decreasing = TRUE) %>%
  print()

# Also check the raw probability rasters -- are they reasonable?
cat("\nMean probability per series across all pixels:\n")
for (s in fitted_series) {
  layer_nm <- paste0("prob_", s)
  if (!layer_nm %in% names(prob_rasters)) next
  vals <- terra::values(prob_rasters[[layer_nm]])
  cat(" ", s, "| mean:", round(mean(vals, na.rm = TRUE), 3),
      "| max:", round(max(vals, na.rm = TRUE), 3), "\n")
}
```


